---
title: "250430Yil - Prerank GSEA processing and visualization - IL22 Experiment"
author: "Charlie Whittaker"
date: "10/7/2025"
output: 
  html_document:
    toc: true
    toc_depth: 4
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, cache=TRUE,
                      dev="png", message=FALSE, error=FALSE, warning=TRUE)
```


# R processing of gene expression data

## Load libraries

```{r, warning=FALSE,error=FALSE,message=FALSE}
library(openxlsx)
library(tidyverse)
library(reprex)
library(matrixStats)
library(XML)
library(ggrepel)
library(rtracklayer)
library(DESeq2)
library(apeglm)
library(ComplexHeatmap)
library(tximport)
library(edgeR)
library(gprofiler2)
library(cluster)
library(stringr)
library(fgsea)
library(UpSetR)
library(ComplexUpset)
library(GseaVis)
```

# load IGB libraries

Loading Custom code

data_formatting_tools.R: functions for organizing, combining, and writing data tables
analysis_tools.R: functions for running differential expression and PCA
plots.R: functions for generating figures

setting the scipen option as such makes R display numbers instead of shortening to scientific notation ("0.01" instead of "1e-02")

```{r}
source('Rcode/data_formatting_tools.R')
source('Rcode/analysis_tools.R')
source('Rcode/plots.R')
source('Rcode/common.R')
source('Rcode/ssGSEA.Library.R')

options(scipen = 9)
```

# Prepare Annotation - ensembl primary assembly basic annotation

```{r}
z <- import("/Genomes/mm39_ens113/Mus_musculus.GRCm39.113.gtf")
tx2gene <- as.data.frame(z) %>% select(c('transcript_id','gene_id'))
tx2gene <- unique(tx2gene)
tx2gene <- tx2gene %>% filter(!is.na(transcript_id))

gene2typesym <- as.data.frame(z) %>% select(c('gene_id','gene_biotype','gene_name'))
gene2typesym <- unique(gene2typesym)

#write.xlsx(tx2gene,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="mm39_ens113_tx2gene.xlsx")
#write.xlsx(gene2typesym,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="mm39_ens113_gene2typesym.xlsx")
```

## Import previously saved annotation files

In subsequent runs, necessary annotations can be loaded in this way from
saved files

```{r}
#tx2gene <- read.xlsx("mm39_ens113_tx2gene.xlsx")
#gene2typesym <- read.xlsx("mm39_ens113_gene2typesym.xlsx")
```

# Importing 3pDGE data from star_salmo

## Read in samples and exclude problematic replicates

```{r}
dir <- getwd()
list.files(dir)

samples <- read.xlsx("250430Yil_Samples_Fangtao_Il22.xlsx")
dim(samples)
samples
# Exclude low count samples
# samples <- samples %>% dplyr::filter(ReadCount > 1000000)
# dim(samples)
```

# GSEA - Functional Class Sorting

## Preparing rank files for GSEA

This extracts protein coding genes from assembled data files, adjusts the data and write rnk file in gsea directory.
This code was executed inside of the processing Rmd and is retained here for reference.

```{r}
# system("mkdir -p ../Il22_gsea && chmod 775 ../Il22_gsea")
# system("cp prerank_4.4.0_v2025.1.sh ../Il22_gsea")
# 
# rnkFile <- assembleDat %>% filter(MPG == "Yes" & gene_biotype == "protein_coding") %>% dplyr::select('gene_name',ends_with(".stat"))
# 
# names <- colnames(rnkFile)
# 
# for (i in names[-1]){
#   d<-rnkFile %>%  dplyr::select('gene_name',all_of(i))
#   names(d)[1] <- '#Gene'
#   d<-d[complete.cases(d), ]
#   write.table(d, sep='\t',file=paste0("../Il22_gsea/",i,".rnk"),col.names=TRUE, quote=FALSE, row.names=FALSE)
# }
```

## Process GSEA results to produce a single output file

### Identify files that need to be imported - the basedir will change according to run

```{r}
# Note: do not include trailing / in base_dir
base_dir <- "../Il22_gsea/oct05"

# Get a list of directories in the base directory
sub_dirs <- list.dirs(base_dir, recursive = FALSE)

# Initialize an empty list to store file paths
file_list <- list()

# Loop through each sub-directory
for (sub_dir in sub_dirs) {
  # Get a list of files in the sub-directory that match the pattern
  files <- list.files(path = sub_dir, pattern = "^gsea_.*\\.tsv$", full.names = TRUE)
  # Add the files to the file list
  file_list <- c(file_list, files)
}
```

### Import the GSEA data into a list of lists

```{r}
gsea_results.list <- lapply(file_list, read.delim)
```

### Function to convert filepaths to shortnames

gsea folder information needs to be adjusted for each run

```{r}
# Define a function to process each file path
process_filepath <- function(filepath) {
  # Remove folder info from the front of the file path
  short_filename <- gsub("../Il22_gsea/oct05/", "", filepath)
  # Replace ".rnk." with "_"
  short_filename <- gsub(".stat.rnk.", "_", short_filename)
  # Add "_pos" if "pos" is in the remainder or "_neg" if "neg" in the remainder
  if (grepl("_pos", short_filename)) {
    short_filename <- paste0(short_filename, "_pos")
  } else if (grepl("neg", short_filename)) {
    short_filename <- paste0(short_filename, "_neg")
  }
  short_filename <- gsub("\\.GseaP.*_neg$", "_neg", short_filename)
  short_filename <- gsub("\\.GseaP.*_pos$", "_pos", short_filename)
  return(short_filename)
}
```

### Process the filepaths

```{r}
# Apply the function to each file path in file_list
short_filenames <- lapply(file_list, process_filepath)
# Rename the list of gsea results
names(gsea_results.list) <- short_filenames
```

### extract relevant columns from the dataframes

```{r}
# Define the columns to select
columns_to_select <- c("NAME", "SIZE", "NES", "FDR.q.val")
# Loop through the list of dataframes
gsea_results.list <- lapply(gsea_results.list, function(df) {
  # Select the desired columns using dplyr::select
  df_selected <- dplyr::select(df, all_of(columns_to_select))
  # Return the modified dataframe
  return(df_selected)
})
```

### Add a comparison column containing the name of each results list

```{r}
# Loop through the list of dataframes
for (comparison_name in names(gsea_results.list)) {
  new_name <- gsub("_pos$|_neg$", "", comparison_name)
  # Access the dataframe by its name
  df <- gsea_results.list[[comparison_name]]
  # Add a new column containing the comparison name
  df <- df %>% mutate(Comparison = new_name)
  # Update the dataframe in the list
  gsea_results.list[[comparison_name]] <- df
}
```

### Ensure that the NES values are numeric

```{r}
gsea_results.list <- lapply(gsea_results.list, function(x) {
  x$NES <- as.numeric(x$NES)
  return(x)
})
```

### Assemble the data into a single output file and export the results

```{r}
non_empty_dfs <- gsea_results.list[sapply(gsea_results.list, nrow) > 0]
combined_df <- bind_rows(non_empty_dfs, .id = "ComparisonFull")

# Use mutate and str_extract to create clean 'Comparison' and 'Collection' column
combined_df <- combined_df %>%
  mutate(Collection = str_extract(Comparison, "[^_]+$")) %>%
  mutate(Comparison = str_remove(Comparison, "_m2cp|_m5bp|_mh"))
```

### Pivot GSEA results

```{r}
# Get the maximum SIZE value for each NAME
df_size_max <- combined_df %>%
  group_by(NAME) %>%
  summarize(SIZE = max(SIZE, na.rm = TRUE), .groups = 'drop')

# Ensure no duplicate entries for the same NAME and Comparison
df_summary <- combined_df %>%
  group_by(NAME, Comparison, Collection) %>%
  summarize(NES = mean(NES, na.rm = TRUE), FDR.q.val = mean(FDR.q.val, na.rm = TRUE), .groups = 'drop')
```

# Pivot the dataframe to get NES and FDR.q.val with correct column names

```{r}
pivoted_df <- df_summary %>%
  pivot_wider(names_from = Comparison,
              values_from = c(NES, FDR.q.val),
              names_sep = ".") %>%
  rename_with(~ sub("^(.+?)\\.NES$", "\\1.NES", .), starts_with("NES")) %>%
  rename_with(~ sub("^(.+?)\\.FDR.q.val$", "\\1.FDR.q.val", .), starts_with("FDR.q.val"))
```

# Combine the SIZE column with the pivoted dataframe

```{r}
final_df <- pivoted_df %>%
  left_join(df_size_max, by = "NAME") %>%
  select(NAME, SIZE, Collection, everything()) # Ensure Collection is included and in the right order

# Identify NES columns
nes_columns <- grep("^NES", names(final_df), value = TRUE)

# Identify FDR.q.val columns
fdr_columns <- grep("^FDR.q.val", names(final_df), value = TRUE)

# Create MaxNES and MinFDR column with the minimum value of all FDR.q.val columns
final_df <- final_df %>%
  rowwise() %>%
  mutate(MaxNES = max(c_across(all_of(nes_columns)), na.rm = TRUE)) %>%
  mutate(MinFDR = min(c_across(all_of(fdr_columns)), na.rm = TRUE)) %>%
  ungroup()

# Sort final_df by MinFDR with smallest values at the top
final_df <- final_df %>% arrange(MinFDR)

# Save the final dataframe
#write.xlsx(final_df, "GSEA_pivot.xlsx")
```

### Generate NES heatmaps

```{r fig.width=4,fig.height=6}
gsea_toPlot <- final_df %>% 
  dplyr::filter(Collection == "mh", MinFDR < 0.05, MaxNES > 2) %>%
  dplyr::select(NAME,starts_with("NES")) %>%
  column_to_rownames(var = "NAME")

# Remove "GOBP_" from the row names
# rownames(gsea_toPlot) <- gsub("GOBP_", "", rownames(gsea_toPlot))

# Create the heatmap with reduced legend size and modified row labels
heatmap <- Heatmap(as.matrix(gsea_toPlot),
                   name = "NES",
                   cluster_rows = TRUE,
                   cluster_columns = TRUE,
                   show_row_names = TRUE,
                   show_column_names = TRUE,
                   row_names_gp = gpar(fontsize = 5),  # Adjust font size for readability
                   column_names_gp = gpar(fontsize = 8), # Adjust font size for readability
                   heatmap_legend_param = list(
                     title_gp = gpar(fontsize = 8),  # Title font size
                     labels_gp = gpar(fontsize = 6),  # Labels font size
                     grid_width = unit(2, "mm"),      # Width of the legend color bar
                     grid_height = unit(2, "mm")      # Height of the legend color bar
                   ))

heatmap
```

## GSEA enrichment plots

### Hallmark Set
Random example genes highlighted

```{r,fig.width=7,fig.height=5}
gene <- c("Ifitm1","Ifitm3","Casp1") 
tt.plot <- gseaNb(object = NULL,
       filePath = "../Il22_gsea/oct05/IL22_vs_WT.stat.rnk.mh.GseaPreranked.1759696017912/",
       geneSetID = "HALLMARK_INTERFERON_ALPHA_RESPONSE",
       subPlot = 2,
       rmHt = F,
       legend.position = c(0.8,0.8),
       termWidth = 50,
       addGene = gene,
       addPval = T,
       pvalX = 0.8,pvalY = 0.8)

tt.plot

#ggsave("gsea_example.pdf", plot = tt.plot, device = "pdf", width = 7, height = 5, dpi = 300)
```

# write session info

Capturing information about the R session helps document your work

```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), "250530Yil_Fangtao_Il22_GSEA_sessionInfo.txt")
```