---
title: "250430Yil - Custom Analyses and Plotting - IL22 Experiment"
author: "Charlie Whittaker"
date: "10/7/2025"
output: 
  html_document:
    toc: true
    toc_depth: 4
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, cache=TRUE,
                      dev="png", message=FALSE, error=FALSE, warning=TRUE)
```


# R processing of gene expression data

## Load libraries

```{r, warning=FALSE,error=FALSE,message=FALSE}
library(openxlsx)
library(tidyverse)
library(reprex)
library(matrixStats)
library(XML)
library(ggrepel)
library(rtracklayer)
library(DESeq2)
library(apeglm)
library(ComplexHeatmap)
library(tximport)
library(edgeR)
library(gprofiler2)
library(cluster)
library(stringr)
library(fgsea)
library(UpSetR)
library(ComplexUpset)
library(GseaVis)
```

# load IGB libraries

Loading Custom code

data_formatting_tools.R: functions for organizing, combining, and writing data tables
analysis_tools.R: functions for running differential expression and PCA
plots.R: functions for generating figures

setting the scipen option as such makes R display numbers instead of shortening to scientific notation ("0.01" instead of "1e-02")

```{r}
source('Rcode/data_formatting_tools.R')
source('Rcode/analysis_tools.R')
source('Rcode/plots.R')
source('Rcode/common.R')
source('Rcode/ssGSEA.Library.R')

options(scipen = 9)
```

# Read in samples and exclude problematic replicates

```{r}
dir <- getwd()
list.files(dir)

samples <- read.xlsx("250430Yil_Samples_Fangtao_Il22.xlsx")
dim(samples)
samples
# Exclude low count samples
# samples <- samples %>% dplyr::filter(ReadCount > 1000000)
# dim(samples)
```

# Read Assembled Data

```{r}
assembleDat <- read.xlsx("250430Yil_Fangtao_Il22_assembled.xlsx")
```

# Upset plots

```{r fig.width=8,fig.height=6}
# Filter the data frame based on your initial criteria
filtered_data <- assembleDat %>% filter(MPG == "Yes" & gene_biotype == "protein_coding")

# Get the column names ending with .padj and .absFC
padj_columns <- names(filtered_data)[str_detect(names(filtered_data), "\\.padj$")]
absfc_columns <- names(filtered_data)[str_detect(names(filtered_data), "\\.absFC$")]

# Extract prefixes for naming the lists
prefixes <- str_remove(padj_columns, "\\.padj$")

# Create a list of data frames, each containing genes for a specific .padj column and .absFC column
gene_lists <- map2(padj_columns, prefixes, ~ {
  list_name <- .y
  fc_column <- paste0(list_name, ".absFC")
  genes <- filtered_data %>%
    filter(.data[[.x]] < 0.05 & .data[[fc_column]] > 1) %>%
    select(gene_name) %>%
    pull(gene_name)
  setNames(list(genes), list_name)
})

# Flatten the list of lists
gene_lists <- flatten(gene_lists)

# Select a substet of the lists

#gene_lists_subset <- gene_lists[c(4,5,7,8)]

# Convert the list of genes into a data frame suitable for UpSet plot
gene_lists_binary <- fromList(gene_lists)

# Generate the UpSet plot
UpSetR::upset(gene_lists_binary, sets = names(gene_lists), order.by = "freq",text.scale = 2,show.numbers = "yes")
```

## Volcano plots

### Prepare gene expression data for plotting

```{r}
vol_df <- assembleDat %>% dplyr::filter(MPG == "Yes" &
                                    gene_biotype == "protein_coding" &
                                    LowExpLowVar == "No") %>% dplyr::select(gene_name, 
                                                                            ends_with(".log2FoldChange"),
                                                                            ends_with(".padj"),
                                                                            ends_with(".absFC"))
```

## Adjust the data table for better plotting

### Replaces NA value padj columns with 1 and NA value log2FoldChange with 0 and sets y axis according to max(abs(LFC))

```{r}
vol_df.edit <- vol_df

vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(is.na(.), 1, .))
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".log2FoldChange")), ~ ifelse(is.na(.), 0, .))
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".absFC")), ~ ifelse(is.na(.), 0, .))
```

### joins list of annotated genes for coloring

```{r}
#vol_df.edit <- merge(vol_df.edit,OfInterest,by.x="gene",by.y="Gene", all.x=TRUE, all.y = FALSE)
#vol_df.edit <- vol_df.edit %>% arrange(!is.na(Pathway),Pathway)
```

### applies a ceiling to -log10(adp)

This may not be necessary but can help when there are a handful of genes with large -log10(adp)

```{r}
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(. < 1e-10, 1e-10, .))
```

### sets y axis according to max(abs(LFC))

```{r}
maxFCs <- vol_df.edit %>% summarize(across(ends_with(".log2FoldChange"), ~ max(abs(.), na.rm = TRUE)))
maxFCs
max(maxFCs)
```

### Prepare vector of plot comparisons

```{r}
comparisons_to_plot <- colnames(vol_df.edit)
comparisons_to_plot <- grep("log2FoldChange$", comparisons_to_plot, value = TRUE)
comparisons_to_plot <- sub("\\.log2FoldChange$", "", comparisons_to_plot)
comparisons_to_plot
```

### Plotting the Comparisons

```{r,fig.width=10,fig.height=10}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  vol_df.edit <- vol_df.edit %>%
    mutate(
      toColor = if_else(vol_df.edit[[col_absFC]] >= 1 & vol_df.edit[[col_padj]] <= 0.05, "Gene", "Empty")
    )

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)
  
  # Create sorted_df so that labels are drawn on top
  # By pvalue
  # sorted_df <- vol_df.edit %>% arrange(vol_df.edit[[col_padj]])
  # By foldChange
  sorted_df <- vol_df.edit %>% arrange(desc(vol_df.edit[[col_absFC]]))

  # Calculate the top n column - adjust the number as desired - these will be labeled with symbol
  sorted_df <- sorted_df %>%
    mutate(top = if_else(row_number() <= 25, gene_name, ""))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for", i)
  p <- ggplot(sorted_df, aes(x = -log10(.data[[col_padj]]), y = .data[[col_log2FoldChange]])) +
    geom_point(aes(color = toColor), size = 2) +
    geom_label_repel(data = filter(sorted_df, !is.na(top)), aes(label = top), box.padding = 0.5, size = 4,max.overlaps=50) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-1, maxFC+1)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)
  print(p)
}
```

# Single sample Gene Set Enrichment Analysis (ssGSEA)

## Import ssGSEA results, row-center the results and export to excel

These data were calculated in the processing Rmd

```{r}
ssGSEA_file_list <- list.files(pattern = "_Il22_ssGSEA.gct$", full.names = TRUE)

read_gct_file <- function(file) {
  read.table(file, header = TRUE, sep = "\t", skip = 2, stringsAsFactors = FALSE)
}

ssGSEA_gct_list <- lapply(ssGSEA_file_list, read_gct_file)
ssGSEA.results <- do.call(rbind, ssGSEA_gct_list)
ssGSEA.results <- ssGSEA.results %>% dplyr::select(-Description)

# filter out "combined up and down gene sets"

#ssGSEA.results <- ssGSEA.results %>% dplyr::filter(!grepl("combined UP & DN", Description))
ssGSEA.results <- ssGSEA.results %>% rename_with(~ str_replace(., "\\.normCt$", ".ssgsea"), ends_with(".normCt"))

ssGSEA.results <- ssGSEA.results %>% remove_rownames %>% column_to_rownames(var="Name")
ssgsea_df <- ssGSEA.results

# Extract group labels from column names - the text pattern here will vary a lot between experiments
sample_info <- tibble(
  Sample = colnames(ssgsea_df),
  Group = str_extract(colnames(ssgsea_df), "(IL22\\.KO|IL22|WT)")
)


# Transpose the data so pathways are rows and samples are columns
ssgsea_df_t <- t(ssgsea_df)
colnames(ssgsea_df_t) <- rownames(ssGSEA.results)
ssgsea_df_t <- as.data.frame(ssgsea_df_t)


# Add group info to the transposed data
ssgsea_df_t$Group <- sample_info$Group
```

# Function to perform t-tests across groups for each pathway

```{r}
test_pathways <- function(data, group_col = "Group") {
  pathways <- colnames(data)[!colnames(data) %in% group_col]
  results <- lapply(pathways, function(pathway) {
    df <- data[, c(pathway, group_col)]
    colnames(df) <- c("Score", "Group")
    
    # Perform ANOVA across all groups
    aov_result <- aov(Score ~ Group, data = df)
    p_val <- summary(aov_result)[[1]][["Pr(>F)"]][1]
    
    # Return pathway name and p-value
    tibble(Pathway = pathway, P_value = p_val)
  })
  
  bind_rows(results) %>% arrange(P_value)
}
```

### Run the test to identify group-level pathway changes

```{r}
pathway_stats <- test_pathways(ssgsea_df_t)
# Select the top n pathways with most significant group-level differences
PathToPlot <- head(pathway_stats$Pathway, 40)
```

## Standardize the data

```{r}
ssGSEA.results <- ssGSEA.results %>% data.matrix()
rc.ssGSEA.results <- calculate_zscore(ssGSEA.results)
rc.ssGSEA.results.df <- as.data.frame(rc.ssGSEA.results)
rc.ssGSEA.results.df <- rc.ssGSEA.results.df %>% rename_with(~ str_replace(., "\\.ssgsea$", ".rc_ssgsea"), ends_with(".ssgsea"))

#write.xlsx(rc.ssGSEA.results.df,colNames=TRUE, quote=FALSE, rowNames=TRUE, file="rc.ssGSEA.results.xlsx")
```

## Heatmap of specific gene sets from ssGSEA

```{r,fig.width=12,fig.height=12}
rc.ssGSEA.results.select <- rc.ssGSEA.results.df %>%
  rownames_to_column("Pathway") %>%
  filter(Pathway %in% PathToPlot) %>%
  column_to_rownames("Pathway")

colnames(rc.ssGSEA.results.select) <- str_remove(colnames(rc.ssGSEA.results.select), "\\.rc_ssgsea$")

# Extract group labels from column names
str_extract(colnames(rc.ssGSEA.results.select), "(IL22\\.KO|IL22|WT)")
group_labels <- str_extract(colnames(rc.ssGSEA.results.select), "(IL22\\.KO|IL22|WT)")

# Ensure it's a proper factor
group_factor <- factor(group_labels, levels = unique(group_labels))

# Define colors for each group
group_colors <- structure(
  c("steelblue", "darkgreen", "salmon"),
  names = levels(group_factor)
)

# Create column annotation (no need to name the factor)
col_ha <- HeatmapAnnotation(
  Group = group_factor,
  col = list(Group = group_colors),
  show_annotation_name = TRUE
)

Heatmap(
  rc.ssGSEA.results.select,
  name = "Z-score",
  top_annotation = col_ha,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8),
  row_names_gp = gpar(fontsize = 8)
)
```

# Heatmaps

## Prepare gene expression data for plotting

```{r}
heat_df <- assembleDat %>% dplyr::filter(MPG == "Yes" &
                                    gene_biotype == "protein_coding" &
                                    LowExpLowVar == "No")

# Reorder data columns according to levels
order.tmp <- samples %>% arrange(Condition)
colOrder <- paste0(order.tmp$Name, ".l2cpm")

heat_df <- heat_df %>% dplyr::select(gene_name, all_of(colOrder),
                                     ends_with(".log2FoldChange"),
                                     ends_with(".padj"),
                                     ends_with(".absFC"))
```

## Adjust the data table for better plotting

Replaces NA value padj columns with 1 and NA value log2FoldChange or absFC with 0 and sets y axis according to max(absLFC)

```{r}
heat_df.edit <- heat_df

heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".log2FoldChange")), ~ ifelse(is.na(.), 0, .))
heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(is.na(.), 1, .))
heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".absFC")), ~ ifelse(is.na(.), 0, .))
```

## Prepare vector of comparisons to plot

```{r}
comparisons_to_plot <- colnames(heat_df.edit)
comparisons_to_plot <- grep("log2FoldChange$", comparisons_to_plot, value = TRUE)
comparisons_to_plot <- sub("\\.log2FoldChange$", "", comparisons_to_plot)
comparisons_to_plot
```

This extracts l2cpm data for some of the most differential expressed
protein coding genes Adjust filters and check row counts

### Plotting the Comparisons - top n fold changes

```{r,fig.width=13,fig.height=10}
for (i in comparisons_to_plot) {
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  heat_df.toPlot <- heat_df.edit %>%
    filter(.data[[col_absFC]] >= 1 & .data[[col_padj]] <= 0.05) %>%
    slice_max(order_by = .data[[col_absFC]], n = 40) %>%
    select(gene_name, ends_with(".l2cpm"))

  # Skip if no rows to plot
  if (nrow(heat_df.toPlot) == 0) {
    message(paste("No DEGs found for", i))
    next
  }

  # Set gene_name as rownames and rename columns
  rownames(heat_df.toPlot) <- heat_df.toPlot$gene_name
  heat_df.toPlot <- heat_df.toPlot %>% select(-gene_name)
  heat_df.toPlot <- heat_df.toPlot %>% rename_with(~ str_remove(., "\\.l2cpm$"))

  # Convert to matrix
  heat_matrix <- as.matrix(heat_df.toPlot)

  # Standardize the data
  heat_matrix.zscore <- calculate_zscore(heat_matrix)
  
  # top annotation
  # Extract group labels from column names
  group_labels <- str_extract(colnames(rc.ssGSEA.results.select), "(IL22\\.KO|IL22|WT)")
  
  # Ensure it's a proper factor
  group_factor <- factor(group_labels, levels = unique(group_labels))
  
  # Define colors for each group
  group_colors <- structure(
    c("steelblue", "darkgreen", "salmon"),
    names = levels(group_factor)
  )
  
  # Create column annotation (no need to name the factor)
  col_ha <- HeatmapAnnotation(
    Group = group_factor,
    col = list(Group = group_colors),
    show_annotation_name = TRUE
  )

  # Create the heatmap
  p <- Heatmap(
    heat_matrix.zscore,
    top_annotation = col_ha,
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    row_names_side = "left",
    column_names_rot = 90,
    column_names_side = "top",
    heatmap_legend_param = list(title = paste0(i, " DEG heatmap"))
  )
  print(p)
}
```

# write session info

Capturing information about the R session helps document your work

```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), "03_250430Yil_Fangtao_IL22_Custom_sessionInfo.txt")
```