---
title: "250812Yil - 3' DGE R processing and analysis - Progenitor Cells"
author: "Charlie Whittaker"
date: "9/30/2025"
output: 
  html_document:
    toc: true
    toc_depth: 4
  html_notebook: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, cache=TRUE,
                      dev="png", message=FALSE, error=FALSE, warning=TRUE)
```

# R processing of gene expression data

## Load libraries

```{r, warning=FALSE,error=FALSE,message=FALSE}
library(openxlsx)
library(tidyverse)
library(reprex)
library(matrixStats)
library(XML)
library(ggrepel)
library(rtracklayer)
library(DESeq2)
library(apeglm)
library(ComplexHeatmap)
library(tximport)
library(edgeR)
library(gprofiler2)
library(cluster)
library(stringr)
library(fgsea)
library(UpSetR)
library(ComplexUpset)
library(GseaVis)
```

# load IGB libraries

Loading Custom code

data_formatting_tools.R: functions for organizing, combining, and
writing data tables analysis_tools.R: functions for running differential
expression and PCA plots.R: functions for generating figures

setting the scipen option as such makes R display numbers instead of
shortening to scientific notation ("0.01" instead of "1e-02")

```{r}
source('Rcode/data_formatting_tools.R')
source('Rcode/analysis_tools.R')
source('Rcode/plots.R')
source('Rcode/common.R')
source('Rcode/ssGSEA.Library.R')

options(scipen = 9)
```

# Prepare Annotation - ensembl primary assembly basic annotation

```{r}
# z <- import("https://ftp.ensembl.org/pub/release-113/gtf/mus_musculus/Mus_musculus.GRCm39.113.gtf.gz")
# tx2gene <- as.data.frame(z) %>% select(c('transcript_id','gene_id'))
# tx2gene <- unique(tx2gene)
# tx2gene <- tx2gene %>% filter(!is.na(transcript_id))
# 
# gene2typesym <- as.data.frame(z) %>% select(c('gene_id','gene_biotype','gene_name'))
# gene2typesym <- unique(gene2typesym)
# 
# write.xlsx(tx2gene,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="mm39_ens113_tx2gene.xlsx")
# write.xlsx(gene2typesym,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="mm39_ens113_gene2typesym.xlsx")
```

## Import previously saved annotation files

In subsequent runs, necessary annotations can be loaded in this way from
saved files

```{r}
tx2gene <- read.xlsx("mm39_ens113_tx2gene.xlsx")
gene2typesym <- read.xlsx("mm39_ens113_gene2typesym.xlsx")
```

# Importing 3pDGE data from star_salmon

## Read in samples

```{r}
dir <- getwd()
#list.files(dir)

samples <- read.xlsx("250812Yil_samples.xlsx")
# Exclude low count samples
# samples <- samples %>% dplyr::filter(ReadCount > 1000000)
dim(samples)
# Select stem samples
samples <- samples %>% dplyr::filter(Cell == "Prog")
dim(samples)
colnames(samples)
table(samples$Condition)
as.data.frame(table(samples$Condition.Simple))
```

```{r}
files <- file.path(dir,samples$Folder, "quant.sf")
files

names(files) <- paste0(samples$Name)
all(file.exists(files))

txi <- tximport(files, type = "salmon", tx2gene = tx2gene)
names(txi)

intCt <- round((txi$counts),0)
colnames(intCt)<-paste0(colnames(intCt),".intCt")

y <- DGEList(txi$counts)
#For creating a matrix of CPMs within edgeR, the following code chunk can be used:
cpms <- cpm(y)
l2cpm <- log((cpms+1),2)
colnames(l2cpm)<-paste0(colnames(l2cpm),".l2cpm")

dumpDat <- merge(intCt, l2cpm, by=0, all=TRUE)
```

## Boxplot of all log2 tpm data to check quality

This plot includes all transcripts/genes in the annotation target.

```{r, fig.width=10,fig.height=6}
dim(l2cpm)

ggplot(stack(as.data.frame(l2cpm)), aes(x = ind, y = values)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("l2cpm boxplot") 
```

# Annotation of gene expression data and post-processing

Once quantified, the gene level salmon data needs to be associated with
additional information to facilitate interpretation

## Annotate assembled data

Note: The annotation file may contain a few examples of multiple
gene/type annotations associated with the same EnsID. This can duplicate
rows in dumpData during merge, check dimensions of objects.

```{r}
assembleDat <- merge(x=dumpDat,y=gene2typesym, by.x=c("Row.names"), by.y=c("gene_id"), all.x=TRUE, all.y = FALSE)

assembleDat <- mutate(assembleDat, Avg = rowMeans(select(assembleDat, ends_with(".l2cpm")), na.rm = TRUE))
assembleDat <- mutate(assembleDat, Var = rowVars(as.matrix(select(assembleDat, ends_with(".l2cpm")), na.rm = TRUE)))
assembleDat <- assembleDat %>% mutate(LowExpLowVar = ifelse(Avg <= 0.1 & Var <= 0.1,"Yes","No"))
```

## Adding MPG column

Often multiple geneids will map to the same gene symbol. In these cases,
it is helpful to select a representative geneID for each group of gene
symbols. The code block below iterates through each unique gene symbol,
compares all of the geneids associated with the same gene symbol and
identifies the geneid with the highest average expression. This gene is
then set as the representative (max-per-gene) geneid for that symbol.
Note: If multiple geneids have the same highest avg expression, the
first gene that appears in the dataframe is taken

The code below performs the following steps: 1) Create a vector called
"mpg" of "No" values the same length as the number of rows in the
dataframe assembleData 2) Loop over the assembleDat object ...
identifying index values of duplicate genes, updating mpg vector 3)
Merge mpg vector to assemblDat object with setting column name as "MPG"

```{r}
mpg <- rep('No', nrow(assembleDat))

for (i in unique(assembleDat[,"gene_name"])){
    if (is.na(i)){next}
    idxs <- which(assembleDat[,"gene_name"] == i)
    max_idx <- which(assembleDat$Avg[idxs] == max(assembleDat$Avg[idxs]))[1]
    
    mpg[idxs[max_idx]] <- 'Yes'
}

assembleDat$MPG <- mpg
```

### Additional boxplots after data filtering

This plot uses the annotation and expression level calculations to
select only expressed protein coding genes

```{r, fig.width=10,fig.height=6}
l2cpm.select <- assembleDat %>% filter(MPG == "Yes" &
                                   gene_biotype == "protein_coding" &
                                   LowExpLowVar == "No") %>% select(ends_with(".l2cpm"))


dim(l2cpm.select)

ggplot(stack(as.data.frame(l2cpm.select)), aes(x = ind, y = values)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("l2cpm Expressed Protein coding boxplot") 
```

## Differential Expression Testing

### Convert columns to factors

```{r}
samples$Condition <- factor(samples$Condition, levels = c("WT_AL_Prog", "WT_Fast_Prog", 
                                                          "Tfe3_AL_Prog", "Tfe3_Fast_Prog", 
                                                          "Tfeb_AL_Prog", "Tfeb_Fast_Prog",
                                                          "Tfe3.Tfeb_AL_Prog","Tfe3.Tfeb_Fast_Prog"))

samples$Condition.Simple <- factor(samples$Condition.Simple, levels = c("WT_AL", "WT_Fast", 
                                                          "Tfe3_AL", "Tfe3_Fast", 
                                                          "Tfeb_AL", "Tfeb_Fast",
                                                          "Tfe3.Tfeb_AL","Tfe3.Tfeb_Fast"))

samples$Animal <- factor(samples$Animal)
samples$Genotype <- factor(samples$Genotype, levels = c("WT","Tfe3","Tfeb","Tfe3.Tfeb"))
samples$Diet <- factor(samples$Diet, levels = c("AL","Fast"))
samples$Cell <- factor(samples$Cell)
samples$GFP_level <- factor(samples$GFP_level)
samples$Gender <- factor(samples$Gender, levels = c("F","M"))
```

### Create dds object from tximport data setting one of the conditions as reference

```{r}
dds <- DESeqDataSetFromMatrix(countData = intCt,
                              colData=samples,
                              design = ~ Condition.Simple)

dds$Condition.Simple <- relevel(dds$Condition.Simple, "WT_AL")
```

## Prepare an exploratory PCA plots

By default, plotPCA using the 500 most variable genes for PCA 1 vs PCA 2
plots this can be adjusted with ntop argument

```{r,fig.width=6,fig.height=6}
vsd <- vst(dds, blind=TRUE)

plotPCA(vsd, intgroup="Name", ntop=500) + geom_point(size = 4) + ggtitle("NameForProcessing")
plotPCA(vsd, intgroup="Condition.Simple", ntop=500) + geom_point(size = 4) + ggtitle("Condition.Simple")
plotPCA(vsd, intgroup="Genotype", ntop=500) + geom_point(size = 4) + ggtitle("Genotype")
plotPCA(vsd, intgroup="Diet", ntop=500) + geom_point(size = 4) + ggtitle("Diet")
plotPCA(vsd, intgroup="Cell", ntop=500) + geom_point(size = 4) + ggtitle("Cell")
plotPCA(vsd, intgroup="Gender", ntop=500) + geom_point(size = 4) + ggtitle("Gender")
plotPCA(vsd, intgroup="ReadCount", ntop=500) + geom_point(size = 4) + ggtitle("ReadCount")
plotPCA(vsd, intgroup="MapRate", ntop=500) + geom_point(size = 4) + ggtitle("Mapping Rate")

pcaData <- plotPCA(vsd, intgroup="Condition.Simple", returnData=TRUE, ntop=500)
pcaData
#write.xlsx(pcaData, colNames=TRUE, rowNames=TRUE, overwrite = TRUE, file="pca.xlsx")
```

### Prepare PCA plot figure

this code needs to be updated to automatically set x and y scales

```{r,fig.height=12,fig.width=14}
# pcaFig <- plotPCA(vsd, intgroup="Condition2", ntop=500) + 
#   geom_point(size = 4) +
#   geom_label_repel(aes(label = Condition2), vjust = -0.5, hjust = 0.5, 
#                    force = 2, min.segment.length = 0.1, max.overlaps = 100) +
#   scale_y_continuous(limits = c(-16, 16), breaks = seq(-16, 16, by = 2)) +
#   scale_x_continuous(limits = c(-22, 22), breaks = seq(-22, 22, by = 2)) +
#   ggtitle("Condition2")
# pcaFig
# 
# pdf('prog_pcaFig.pdf',width=12, height=14)
# pcaFig
# dev.off()
```

# Check counts for some genes of interest and gender genes

Adjust the list as needed

```{r}
genes.To.check <- c("Tfe3", "Tfeb", "Lgr5", "Ddx3y", "Xist")

for (gene in genes.To.check) {
  ens.ID <- gene2typesym %>% filter(gene_name == gene)
  
  if (nrow(ens.ID) == 0) {
    warning(paste("Gene", gene, "not found in gene2typesym"))
    next
  }

  cts.ToPlot <- plotCounts(dds, gene = ens.ID$gene_id, intgroup = "Condition.Simple", returnData = TRUE)
  #print(cts.ToPlot)
  
  p <- ggplot(cts.ToPlot, aes(x = Condition.Simple, y = count)) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.9, color = "brown4") +
    labs(title = paste("Normalized counts for", gene), y = "Normalized count")
  
  print(p)
}
```

# Differential Expression Comparisons

## There are 28 distinct comparisons:

### Reference: WT_AL

-   WT_Fast vs WT_AL\
-   Tfe3_AL vs WT_AL\
-   Tfe3_Fast vs WT_AL\
-   Tfeb_AL vs WT_AL\
-   Tfeb_Fast vs WT_AL\
-   Tfe3.Tfeb_AL vs WT_AL\
-   Tfe3.Tfeb_Fast vs WT_AL

### Reference: WT_Fast

-   Tfe3_AL vs WT_Fast\
-   Tfe3_Fast vs WT_Fast\
-   Tfeb_AL vs WT_Fast\
-   Tfeb_Fast vs WT_Fast\
-   Tfe3.Tfeb_AL vs WT_Fast\
-   Tfe3.Tfeb_Fast vs WT_Fast

### Reference: Tfe3_AL

-   Tfe3_Fast vs Tfe3_AL\
-   Tfeb_AL vs Tfe3_AL\
-   Tfeb_Fast vs Tfe3_AL\
-   Tfe3.Tfeb_AL vs Tfe3_AL\
-   Tfe3.Tfeb_Fast vs Tfe3_AL

### Reference: Tfe3_Fast

-   Tfeb_AL vs Tfe3_Fast\
-   Tfeb_Fast vs Tfe3_Fast\
-   Tfe3.Tfeb_AL vs Tfe3_Fast\
-   Tfe3.Tfeb_Fast vs Tfe3_Fast

### Reference: Tfeb_AL

-   Tfeb_Fast vs Tfeb_AL\
-   Tfe3.Tfeb_AL vs Tfeb_AL\
-   Tfe3.Tfeb_Fast vs Tfeb_AL

### Reference: Tfeb_Fast

-   Tfe3.Tfeb_AL vs Tfeb_Fast\
-   Tfe3.Tfeb_Fast vs Tfeb_Fast

### Reference: Tfe3.Tfeb_AL

-   Tfe3.Tfeb_Fast vs Tfe3.Tfeb_AL

## Run tests vs WT_AL

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs WT_AL_stem comparisons coef

```{r}
vs_WT_AL.comparisons <- resultsNames(dds)[c(2,3,4,5,6,7,8)]
vs_WT_AL.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_WT_AL.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition.Simple_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to WT_Fast

```{r}
dds$Condition.Simple <- relevel(dds$Condition.Simple, "WT_Fast")
```

### Run tests vs WT_Fast

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs F.WT_stem comparisons coef

```{r}
vs_WT_Fast.comparisons <- resultsNames(dds)[c(3,4,5,6,7,8)]
vs_WT_Fast.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_WT_Fast.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition.Simple_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to Rf2h.WT_stem

```{r}
dds$Condition.Simple <- relevel(dds$Condition.Simple, "Tfe3_AL")
```

### Run tests vs Rf2h.WT_stem

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs Rf2h.WT_stem comparisons coef

```{r}
vs_Tfe3_AL.comparisons <- resultsNames(dds)[c(4,5,6,7,8)]
vs_Tfe3_AL.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_Tfe3_AL.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition.Simple_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to Tfe3_Fast

```{r}
dds$Condition.Simple <- relevel(dds$Condition.Simple, "Tfe3_Fast")
```

### Run tests vs Tfe3_Fast

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs Tfe3_Fast comparisons coef

```{r}
vs_Tfe3_Fast.comparisons <- resultsNames(dds)[c(5,6,7,8)]
vs_Tfe3_Fast.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_Tfe3_Fast.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition.Simple_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to Tfeb_AL

```{r}
dds$Condition.Simple <- relevel(dds$Condition.Simple, "Tfeb_AL")
```

### Run tests vs Tfeb_AL

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs Tfeb_AL comparisons coef

```{r}
vs_Tfeb_AL.comparisons <- resultsNames(dds)[c(6,7,8)]
vs_Tfeb_AL.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_Tfeb_AL.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition.Simple_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to Tfeb_Fast

```{r}
dds$Condition.Simple <- relevel(dds$Condition.Simple, "Tfeb_Fast")
```

### Run tests vs Tfeb_Fast

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs Tfeb_Fast comparisons coef

```{r}
vs_Tfeb_Fast.comparisons <- resultsNames(dds)[c(7,8)]
vs_Tfeb_Fast.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_Tfeb_Fast.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition.Simple_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to Tfe3.Tfeb_AL

```{r}
dds$Condition.Simple <- relevel(dds$Condition.Simple, "Tfe3.Tfeb_AL")
```

### Run tests vs Tfe3.Tfeb_AL

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs Tfe3.Tfeb_AL comparisons coef

```{r}
vs_Tfe3.Tfeb_AL.comparisons <- resultsNames(dds)[c(8)]
vs_Tfe3.Tfeb_AL.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_Tfe3.Tfeb_AL.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition.Simple_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Write output file

```{r}
write.xlsx(assembleDat,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="250812Yil_prog_assembled.xlsx")
```

# Comparison list for checking

```{r}
# ls(pattern=".comparisons")

# Combine all comparisons in order
comparisons.check <- data.frame(
  comparisons = c(
    vs_WT_AL.comparisons,
    vs_WT_Fast.comparisons,
    vs_Tfe3_AL.comparisons,
    vs_Tfe3_Fast.comparisons,
    vs_Tfeb_AL.comparisons,
    vs_Tfeb_Fast.comparisons,
    vs_Tfe3.Tfeb_AL.comparisons
  )
)

# View the result
comparisons.check
```

## Upset Plots

```{r fig.width=8,fig.height=6}
# Filter the data frame based on your initial criteria
filtered_data <- assembleDat %>% filter(MPG == "Yes" & gene_biotype == "protein_coding")

# Get the column names ending with .padj and .absFC
padj_columns <- names(filtered_data)[str_detect(names(filtered_data), "\\.padj$")]
absfc_columns <- names(filtered_data)[str_detect(names(filtered_data), "\\.absFC$")]

# Extract prefixes for naming the lists
prefixes <- str_remove(padj_columns, "\\.padj$")

# Create a list of data frames, each containing genes for a specific .padj column and .absFC column
gene_lists <- map2(padj_columns, prefixes, ~ {
  list_name <- .y
  fc_column <- paste0(list_name, ".absFC")
  genes <- filtered_data %>%
    filter(.data[[.x]] < 0.05 & .data[[fc_column]] > 1) %>%
    select(gene_name) %>%
    pull(gene_name)
  setNames(list(genes), list_name)
})

# Flatten the list of lists
gene_lists <- flatten(gene_lists)

# Select a substet of the lists

gene_lists_subset <- gene_lists[c(1:7)]
names(gene_lists_subset)

# Convert the list of genes into a data frame suitable for UpSet plot
gene_lists_binary <- fromList(gene_lists_subset)

# Generate the UpSet plot
UpSetR::upset(gene_lists_binary, sets = names(gene_lists_subset), order.by = "freq",text.scale = 2,show.numbers = "yes")
```

# Volcano plots

## Prepare gene expression data for plotting

```{r}
vol_df <- assembleDat %>% dplyr::filter(MPG == "Yes" &
                                    gene_biotype == "protein_coding" &
                                    LowExpLowVar == "No") %>% dplyr::select(gene_name, 
                                                                            ends_with(".log2FoldChange"),
                                                                            ends_with(".padj"),
                                                                            ends_with(".absFC"))
```

## Adjust the data table for better plotting

### Replaces NA value padj columns with 1 and NA value log2FoldChange with 0 and sets y axis according to max(abs(LFC))

```{r}
vol_df.edit <- vol_df

vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".log2FoldChange")), ~ ifelse(is.na(.), 0, .))
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(is.na(.), 1, .))
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".absFC")), ~ ifelse(is.na(.), 0, .))
```

### applies a ceiling to -log10(adp)

This may not be necessary but can help when there are a handful of genes with large -log10(adp)

```{r}
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(. < 1e-10, 1e-10, .))
```

### identify max fold changes to consider y axes in plots

```{r}
maxFCs <- vol_df.edit %>% summarize(across(ends_with(".log2FoldChange"), ~ max(abs(.), na.rm = TRUE)))
maxFCs
max(maxFCs)
```

## Annotating genes for coloring

### Manual creation of a gene list to highlight

```{r}
fasting <- as.data.frame(c("Acaa1a","Acaa1b","Acad8","Acadm","Acot2","Acox1","Acsl3",
                           "Angptl4","Cpt1a","Crot","Decr1","Decr2","Dgat1","Ech1","Etfb",
                           "Fabp1","Fabp2","Hoxa1","Hadhb","Hmgcl","Hmgcs2","Jag2","Pdk4","Ucp2"))

colnames(fasting) <- "gene_name"
fasting <- fasting %>% dplyr::mutate(Fasting = gene_name)
```

### Creating a gene list to highlight from differential expression results

```{r}
upWTfast <- assembleDat %>% dplyr::filter(gene_biotype == "protein_coding",
                                                  MPG == "Yes",
                                                  #str_starts(gene_name, "Hsp"),
                                                  WT_Fast_vs_WT_AL.log2FoldChange > 1,
                                                  WT_Fast_vs_WT_AL.padj < 0.05) %>% dplyr::select(gene_name)

upWTfast <- upWTfast %>% dplyr::mutate(upWTfast = gene_name)
```

### Join example lists to plotting df

```{r}
vol_df.edit <- merge(vol_df.edit,fasting,by.x="gene_name",by.y="gene_name", all.x=TRUE, all.y = FALSE)
vol_df.edit <- merge(vol_df.edit,upWTfast,by.x="gene_name",by.y="gene_name", all.x=TRUE, all.y = FALSE)
```

## Prepare vector of comparisons to plot

```{r}
comparisons_to_plot <- colnames(vol_df.edit)
comparisons_to_plot <- grep("log2FoldChange$", comparisons_to_plot, value = TRUE)
comparisons_to_plot <- sub("\\.log2FoldChange$", "", comparisons_to_plot)
comparisons_to_plot
```

### Plotting the Comparisons - top n fold changes

```{r,fig.width=10,fig.height=10}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  vol_df.edit <- vol_df.edit %>%
    mutate(
      toColor = if_else(vol_df.edit[[col_absFC]] >= 1 & vol_df.edit[[col_padj]] <= 0.05, "Gene", "Empty")
    )

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)
  
  # Create sorted_df so that labels are drawn on top
  # By pvalue
  # sorted_df <- vol_df.edit %>% arrange(vol_df.edit[[col_padj]])
  # By foldChange
  sorted_df <- vol_df.edit %>% arrange(desc(vol_df.edit[[col_absFC]]))

  # Calculate the top n column - adjust the number as desired - these will be labeled with symbol
  sorted_df <- sorted_df %>%
    mutate(top = if_else(row_number() <= 25, gene_name, ""))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for", i)
  p <- ggplot(sorted_df, aes(x = -log10(.data[[col_padj]]), y = .data[[col_log2FoldChange]])) +
    geom_point(aes(color = toColor), size = 2) +
    geom_label_repel(data = filter(sorted_df, !is.na(top)), aes(label = top), box.padding = 0.5, size = 4,max.overlaps=50) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-1, maxFC+1)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)
  print(p)
}
```

### Plotting the Comparisons - gene set of interest

```{r,fig.width=8,fig.height=8}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)

  # Separate data frames for points and labels
  points_df <- sorted_df %>% filter(is.na(Fasting))
  labels_df <- sorted_df %>% filter(!is.na(Fasting))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for fasting genes - ", i)
  p <- ggplot() +
    geom_point(data = points_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "grey70"), size = 2) +
    geom_point(data = labels_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "red"), size = 3) +
    geom_label_repel(data = labels_df, aes(x = -log10(.data[[col_padj]]),
                                           y = .data[[col_log2FoldChange]],
                                           label = Fasting), box.padding = 0.5, size = 3, max.overlaps = 100) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-0.5, maxFC+0.5)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)

  print(p)
}
```


### Plotting the Comparisons - selected genes

```{r,,fig.width=8,fig.height=8}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)

  # Separate data frames for points and labels
  points_df <- sorted_df %>% filter(is.na(upWTfast))
  labels_df <- sorted_df %>% filter(!is.na(upWTfast))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for fasting genes - ", i)
  p <- ggplot() +
    geom_point(data = points_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "grey70"), size = 2) +
    geom_point(data = labels_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "red"), size = 3) +
    geom_label_repel(data = labels_df, aes(x = -log10(.data[[col_padj]]),
                                           y = .data[[col_log2FoldChange]],
                                           label = upWTfast), box.padding = 0.5, size = 3, max.overlaps = 100) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-0.5, maxFC+0.5)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)

  print(p)
}
```

# Check counts for some differentially expressed genes

Adjust the list as needed

```{r}
genes.To.check <- c("Cpt1a", "Fabp1", "Fabp2", "Decr1", "Hmgcs2")

for (gene in genes.To.check) {
  ens.ID <- gene2typesym %>% filter(gene_name == gene)

  if (nrow(ens.ID) == 0) {
    warning(paste("Gene", gene, "not found in gene2typesym"))
    next
  }

  cts.ToPlot <- plotCounts(dds, gene = ens.ID$gene_id, intgroup = "Condition.Simple", returnData = TRUE)
  print(cts.ToPlot)

  cts.ToPlot$Condition.Simple <- factor(cts.ToPlot$Condition.Simple, levels = c("WT_AL","WT_Fast",
                                                                         "Tfe3_AL", "Tfe3_Fast",
                                                                         "Tfeb_AL","Tfeb_Fast",
                                                                         "Tfe3.Tfeb_AL","Tfe3.Tfeb_Fast"))

  p <- ggplot(cts.ToPlot, aes(x = Condition.Simple, y = count)) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.9, color = "brown4") +
    labs(title = paste("Normalized counts for", gene), y = "Normalized count")

  print(p)
}
```

# GSEA - Functional Class Sorting

## Preparing rank files for GSEA

this code chunk prepares rnk files for Linux GSEA runs
Create the directory, extract protein coding genes and write rnk file in gsea directory

```{r}
system("mkdir -p ../gsea && chmod 775 ../gsea")

rnkFile <- assembleDat %>% filter(MPG == "Yes" & gene_biotype == "protein_coding") %>% dplyr::select('gene_name',ends_with(".stat"))

names <- colnames(rnkFile)

for (i in names[-1]){
  d<-rnkFile %>%  dplyr::select('gene_name',all_of(i))
  names(d)[1] <- '#Gene'
  d<-d[complete.cases(d), ]
  write.table(d, sep='\t',file=paste0("../gsea/",i,".prog.rnk"),col.names=TRUE, quote=FALSE, row.names=FALSE)
}
```

# Single sample Gene Set Enrichment Analysis (ssGSEA)

## Extract and prepare gene expression matrix in .gct format from dds object

```{r}
# Extract the normalized count matrix from the dds
normalized_counts <- as.data.frame(counts(dds, normalized = TRUE))
# rename columns - this depends on 3pDGE vs standard
normalized_counts <- normalized_counts %>% rename_with(~ str_replace(., "\\.intCt$", ".normCt"), ends_with(".intCt"))
# if standard RNA-Seq just append .normCt
#normalized_counts <- rename_with(normalized_counts, ~paste0(.x, ".normCt"))

# Annotate counts table with gene information
normalized_counts <- merge(normalized_counts, gene2typesym, by.x=0, by.y="gene_id", all.x=TRUE, all.y=TRUE)

# Annotate counts table with MPG from assembleData
expInfo <- assembleDat %>% dplyr::select(Row.names,MPG)
normalized_counts <- merge(normalized_counts,expInfo, by.x="Row.names", by.y="Row.names", all.x=TRUE, all.y=TRUE)

# Calculate row sum
normalized_counts <- normalized_counts %>% mutate(Sum = rowSums(select(., ends_with(".normCt"))))

# Filter to protein coding genes, MPG = yes and Sum  > 0
normalized_counts <- normalized_counts %>% dplyr::filter(gene_biotype == "protein_coding" & MPG == "Yes" & Sum > 0)

# Reorder dataframe and select only columns needed for .gct file
normalized_counts <- normalized_counts %>% dplyr::select(gene_name,Row.names,ends_with(".normCt"))

# Reorder data columns according to levels
normalized_counts <- normalized_counts %>% dplyr::select(gene_name,Row.names,
                                                         contains("WT_AL_Prog"),contains("WT_Fast_Prog"),
                                                         contains("Tfe3_AL_Prog"),contains("Tfe3_Fast_Prog"),
                                                         contains("Tfeb_AL_Prog"),contains("Tfeb_Fast_Prog"),
                                                         contains("Tfe3.Tfeb_AL_Prog"),contains("Tfe3.Tfeb_Fast_Prog"))


# Rename ID columns
normalized_counts <- normalized_counts %>% dplyr::rename(NAME = gene_name, Description = Row.names)

# Prepare gct file
numDatCols <- normalized_counts %>% select(ends_with(".normCt")) %>% colnames() %>% length()
numDatRows <- normalized_counts %>% rownames() %>% length()

filename1 <- paste("250812Yil_prog_normalizedCounts", ".gct", sep="")
cat("#1.2", "\n", sep="\t", file=filename1)
cat(numDatRows, numDatCols, "\n", sep="\t", file=filename1, append = TRUE)

write.table(normalized_counts, file=filename1, row.names=FALSE, quote = FALSE, sep="\t", append=TRUE)
```

## run ssGSEA

The path on luria to gene sets is:
/net/bmc-lab3/data/bcc/charliew/annotationFiles/v2025/ 
I need to make this path available to the singularity Rstudio
Currently im copying the necessary gmt files into the analysis dir

```{r}
ssGSEA.project.dataset(input.ds = "250812Yil_prog_normalizedCounts.gct", output.ds = "mh_prog_ssGSEA", gene.sets.dbfile.list = "mh.all.v2025.1.Mm.symbols.gmt")
ssGSEA.project.dataset(input.ds = "250812Yil_prog_normalizedCounts.gct", output.ds = "m2cp_prog_ssGSEA", gene.sets.dbfile.list = "m2.cp.v2025.1.Mm.symbols.gmt")
```

## Import ssGSEA results, row-center the results and export to excel

```{r}
ssGSEA_file_list <- list.files(pattern = "ssGSEA.gct$", full.names = TRUE)

read_gct_file <- function(file) {
  read.table(file, header = TRUE, sep = "\t", skip = 2, stringsAsFactors = FALSE)
}

ssGSEA_gct_list <- lapply(ssGSEA_file_list, read_gct_file)
ssGSEA.results <- do.call(rbind, ssGSEA_gct_list)

# filter out "combined up and down gene sets"

#ssGSEA.results <- ssGSEA.results %>% dplyr::filter(!grepl("combined UP & DN", Description))
ssGSEA.results <- ssGSEA.results %>% rename_with(~ str_replace(., "\\.normCt$", ".ssgsea"), ends_with(".normCt"))
ssGSEA.results <- ssGSEA.results %>% select(Name,ends_with(".ssgsea"))

ssGSEA.results <- ssGSEA.results %>% remove_rownames %>% column_to_rownames(var="Name")
ssgsea_df <- ssGSEA.results

# Extract group labels from column names - the text pattern here will vary a lot between experiments
sample_info <- tibble(
  Sample = colnames(ssgsea_df),
  Group = str_extract(colnames(ssgsea_df), "(WT|Tfe3|Tfeb|Tfe3\\.Tfeb)_(AL|Fast)_Prog")
)

# Transpose the data so pathways are rows and samples are columns
ssgsea_df_t <- t(ssgsea_df)
colnames(ssgsea_df_t) <- rownames(ssGSEA.results)
ssgsea_df_t <- as.data.frame(ssgsea_df_t)


# Add group info to the transposed data
ssgsea_df_t$Group <- sample_info$Group
```

# Function to perform t-tests across groups for each pathway

```{r}
test_pathways <- function(data, group_col = "Group") {
  pathways <- colnames(data)[!colnames(data) %in% group_col]
  results <- lapply(pathways, function(pathway) {
    df <- data[, c(pathway, group_col)]
    colnames(df) <- c("Score", "Group")
    
    # Perform ANOVA across all groups
    aov_result <- aov(Score ~ Group, data = df)
    p_val <- summary(aov_result)[[1]][["Pr(>F)"]][1]
    
    # Return pathway name and p-value
    tibble(Pathway = pathway, P_value = p_val)
  })
  
  bind_rows(results) %>% arrange(P_value)
}
```

### Run the test to identify group-level pathway changes

```{r}
pathway_stats <- test_pathways(ssgsea_df_t)
# Select the top n pathways with most significant group-level differences
PathToPlot <- head(pathway_stats$Pathway, 30)
```

## Standardize the data

```{r}
ssGSEA.results <- ssGSEA.results %>% data.matrix()
rc.ssGSEA.results <- calculate_zscore(ssGSEA.results)
rc.ssGSEA.results.df <- as.data.frame(rc.ssGSEA.results)
rc.ssGSEA.results.df <- rc.ssGSEA.results.df %>% rename_with(~ str_replace(., "\\.ssgsea$", ".rc_ssgsea"), ends_with(".ssgsea"))

#write.xlsx(rc.ssGSEA.results.df,colNames=TRUE, quote=FALSE, rowNames=TRUE, file="rc.ssGSEA.results.xlsx")
```

## Heatmap of specific gene sets from ssGSEA

```{r,fig.width=12,fig.height=12}
rc.ssGSEA.results.select <- rc.ssGSEA.results.df %>%
  rownames_to_column("Pathway") %>%
  filter(Pathway %in% PathToPlot) %>%
  column_to_rownames("Pathway")

colnames(rc.ssGSEA.results.select) <- str_remove(colnames(rc.ssGSEA.results.select), "_Prog\\.rc_ssgsea$")

# Extract group labels from column names
group_labels <- str_extract(colnames(rc.ssGSEA.results.select), "(WT|Tfe3|Tfeb|Tfe3\\.Tfeb)_(AL|Fast)")

# Ensure it's a proper factor
group_factor <- factor(group_labels, levels = unique(group_labels))

# Define colors for each group
group_colors <- structure(
  c("steelblue", "skyblue", "darkgreen", "lightgreen", "darkred", "salmon", "purple", "orchid"),
  names = levels(group_factor)
)

# Create column annotation (no need to name the factor)
col_ha <- HeatmapAnnotation(
  Group = group_factor,
  col = list(Group = group_colors),
  show_annotation_name = TRUE
)

Heatmap(
  rc.ssGSEA.results.select,
  name = "Z-score",
  top_annotation = col_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8),
  row_names_gp = gpar(fontsize = 8)
)
```

# Heatmaps

## Prepare gene expression data for plotting

```{r}
heat_df <- assembleDat %>% dplyr::filter(MPG == "Yes" &
                                    gene_biotype == "protein_coding" &
                                    LowExpLowVar == "No") %>% dplyr::select(gene_name,ends_with("WT_AL_Prog.l2cpm"),ends_with("WT_Fast_Prog.l2cpm"),
                                                                            ends_with("Tfe3_AL_Prog.l2cpm"),ends_with("Tfe3_Fast_Prog.l2cpm"),
                                                                            ends_with("Tfeb_AL_Prog.l2cpm"),ends_with("Tfeb_Fast_Prog.l2cpm"),
                                                                            ends_with("Tfe3.Tfeb_AL_Prog.l2cpm"),ends_with("Tfe3.Tfeb_Fast_Prog.l2cpm"),
                                                                            ends_with(".log2FoldChange"),
                                                                            ends_with(".padj"),
                                                                            ends_with(".absFC"))
```

## Adjust the data table for better plotting

### Replaces NA value padj columns with 1 and NA value log2FoldChange with 0 and sets y axis according to max(abs(LFC))

```{r}
heat_df.edit <- heat_df

heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".log2FoldChange")), ~ ifelse(is.na(.), 0, .))
heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(is.na(.), 1, .))
heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".absFC")), ~ ifelse(is.na(.), 0, .))
```

## Prepare vector of comparisons to plot

```{r}
comparisons_to_plot <- colnames(heat_df.edit)
comparisons_to_plot <- grep("log2FoldChange$", comparisons_to_plot, value = TRUE)
comparisons_to_plot <- sub("\\.log2FoldChange$", "", comparisons_to_plot)
comparisons_to_plot
```

This extracts l2cpm data for some of the most differential expressed
protein coding genes Adjust filters and check row counts

### Plotting the Comparisons - top n fold changes

```{r,fig.width=13,fig.height=10}
for (i in comparisons_to_plot) {
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  heat_df.toPlot <- heat_df.edit %>%
    filter(.data[[col_absFC]] >= 1 & .data[[col_padj]] <= 0.05) %>%
    slice_max(order_by = .data[[col_absFC]], n = 40) %>%
    select(gene_name, ends_with(".l2cpm"))

  # Skip if no rows to plot
  if (nrow(heat_df.toPlot) == 0) {
    message(paste("No DEGs found for", i))
    next
  }

  # Set gene_name as rownames and rename columns
  rownames(heat_df.toPlot) <- heat_df.toPlot$gene_name
  heat_df.toPlot <- heat_df.toPlot %>% select(-gene_name)
  heat_df.toPlot <- heat_df.toPlot %>% rename_with(~ str_remove(., "\\.l2cpm$"))

  # Convert to matrix
  heat_matrix <- as.matrix(heat_df.toPlot)

  # Standardize the data
  heat_matrix.zscore <- calculate_zscore(heat_matrix)
  
  # top annotation
  # Extract group labels from column names
  group_labels <- str_extract(colnames(heat_matrix.zscore), "(WT|Tfe3|Tfeb|Tfe3\\.Tfeb)_(AL|Fast)")
  
  # Ensure it's a proper factor
  group_factor <- factor(group_labels, levels = unique(group_labels))
  
  # Define colors for each group
  group_colors <- structure(
    c("steelblue", "skyblue", "darkgreen", "lightgreen", "darkred", "salmon", "purple", "orchid"),
    names = levels(group_factor)
  )
  
  # Create column annotation (no need to name the factor)
  col_ha <- HeatmapAnnotation(
    Group = group_factor,
    col = list(Group = group_colors),
    show_annotation_name = TRUE
  )

  # Create the heatmap
  p <- Heatmap(
    heat_matrix.zscore,
    top_annotation = col_ha,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    row_names_side = "left",
    column_names_rot = 90,
    column_names_side = "top",
    heatmap_legend_param = list(title = paste0(i, " DEG heatmap"))
  )
  print(p)
}
```

# Supplemental data and GEO files

```{r}
system("mkdir -p ../geo_submission && chmod 775 ../geo_submission")
IntCt.geo <- assembleDat %>% dplyr::select(Row.names,ends_with(".intCt"))
IntCt.geo <- IntCt.geo %>% dplyr::rename(ENSID = Row.names)
IntCt.geo <- IntCt.geo %>% dplyr::rename_all(~str_replace(., "\\.intCt", ""))
write.table(IntCt.geo,file="../geo_submission/250812Yil_prog_intCt.txt", sep = "\t", quote = FALSE, row.names = FALSE)

l2cpm.geo <- assembleDat %>% dplyr::select(Row.names,ends_with(".l2cpm"))
l2cpm.geo <- l2cpm.geo %>% dplyr::rename(ENSID = Row.names)
l2cpm.geo <- l2cpm.geo %>% dplyr::rename_all(~str_replace(., "\\.l2cpm", ""))
write.table(l2cpm.geo,file="../geo_submission/250812Yil_prog_l2tpm.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```

# write session info

Capturing information about the R session helps document your work

```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), "250812Yil_prog_sessionInfo.txt")
```
