---
title: "BMCID - 3' DGE R processing and analysis - Details here"
author: "Charlie Whittaker"
date: "xx/xx/20xx"
output: 
  html_document:
    toc: true
    toc_depth: 4
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, cache=TRUE,
                      dev="png", message=FALSE, error=FALSE, warning=TRUE)
```


# R processing of gene expression data

## Load libraries

```{r, warning=FALSE,error=FALSE,message=FALSE}
library(openxlsx)
library(tidyverse)
library(reprex)
library(matrixStats)
library(XML)
library(ggrepel)
library(rtracklayer)
library(DESeq2)
library(apeglm)
library(ComplexHeatmap)
library(tximport)
library(edgeR)
library(gprofiler2)
library(cluster)
library(stringr)
library(fgsea)
library(UpSetR)
library(ComplexUpset)
library(GseaVis)
```

# load IGB libraries

Loading Custom code

data_formatting_tools.R: functions for organizing, combining, and writing data tables
analysis_tools.R: functions for running differential expression and PCA
plots.R: functions for generating figures

setting the scipen option as such makes R display numbers instead of shortening to scientific notation ("0.01" instead of "1e-02")

```{r}
source('Rcode/data_formatting_tools.R')
source('Rcode/analysis_tools.R')
source('Rcode/plots.R')
source('Rcode/common.R')
source('Rcode/ssGSEA.Library.R')

options(scipen = 9)
```

# Prepare Annotation - ensembl primary assembly basic annotation

```{r}
z <- import("/Genomes/mm39_ens113/Mus_musculus.GRCm39.113.gtf")
tx2gene <- as.data.frame(z) %>% select(c('transcript_id','gene_id'))
tx2gene <- unique(tx2gene)
tx2gene <- tx2gene %>% filter(!is.na(transcript_id))

gene2typesym <- as.data.frame(z) %>% select(c('gene_id','gene_biotype','gene_name'))
gene2typesym <- unique(gene2typesym)

#write.xlsx(tx2gene,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="mm39_ens113_tx2gene.xlsx")
#write.xlsx(gene2typesym,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="mm39_ens113_gene2typesym.xlsx")
```

## Import previously saved annotation files

In subsequent runs, necessary annotations can be loaded in this way from saved files

```{r}
#tx2gene <- read.xlsx("mm39_ens113_tx2gene.xlsx")
#gene2typesym <- read.xlsx("mm39_ens113_gene2typesym.xlsx")
```

# Importing 3pDGE data from star_salmon

## Read in samples and exclude problematic replicates

```{r}
dir <- getwd()
list.files(dir)

samples <- read.xlsx("250430Yil_Samples_Shinya.xlsx")
dim(samples)
# Exclude low count samples
samples %>% dplyr::filter(ReadCount < 1000000) %>% dplyr::select(NameForProcessing,Condition,Cell,Diet,Odc1_status,Gender,ReadCount)
samples <- samples %>% dplyr::filter(ReadCount > 1000000)
dim(samples)
# Select stem samples
samples <- samples %>% dplyr::filter(Cell == "Stem")
dim(samples)
samples
```

```{r}
files <- file.path(dir,samples$Folder, "quant.sf")
files

names(files) <- paste0(samples$NameForProcessing)
all(file.exists(files))

txi <- tximport(files, type = "salmon", tx2gene = tx2gene)
names(txi)

intCt <- round((txi$counts),0)
colnames(intCt)<-paste0(colnames(intCt),".intCt")

y <- DGEList(txi$counts)
#For creating a matrix of CPMs within edgeR, the following code chunk can be used:
cpms <- cpm(y)
l2cpm <- log((cpms+1),2)
colnames(l2cpm)<-paste0(colnames(l2cpm),".l2cpm")

dumpDat <- merge(intCt, l2cpm, by=0, all=TRUE)
```

## Boxplot of all log2 tpm data to check quality

This plot includes all transcripts/genes in the annotation target.

```{r, fig.width=10,fig.height=6}
dim(l2cpm)

ggplot(stack(as.data.frame(l2cpm)), aes(x = ind, y = values)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("l2cpm boxplot") 
```

# Annotation of gene expression data and post-processing

Once quantified, the gene level salmon data needs to be associated with additional information to facilitate interpretation

## Annotate assembled data

Note: The annotation file may contain a few examples of multiple gene/type annotations associated with the same EnsID.
This can duplicate rows in dumpData during merge, check dimensions of objects.

```{r}
assembleDat <- merge(x=dumpDat,y=gene2typesym, by.x=c("Row.names"), by.y=c("gene_id"), all.x=TRUE, all.y = FALSE)

assembleDat <- mutate(assembleDat, Avg = rowMeans(select(assembleDat, ends_with(".l2cpm")), na.rm = TRUE))
assembleDat <- mutate(assembleDat, Var = rowVars(as.matrix(select(assembleDat, ends_with(".l2cpm")), na.rm = TRUE)))
assembleDat <- assembleDat %>% mutate(LowExpLowVar = ifelse(Avg <= 0.1 & Var <= 0.1,"Yes","No"))
```

## Adding MPG column

Often multiple geneids will map to the same gene symbol. In these cases, it is helpful to select a representative geneID for each group of gene symbols. The code block below iterates through each unique gene symbol, compares all of the geneids associated with the same gene symbol and identifies the geneid with the highest average expression. This gene is then set as the representative (max-per-gene) geneid for that symbol. Note: If multiple geneids have the same highest avg expression, the first gene that appears in the dataframe is taken

The code below performs the following steps:
  1) Create a vector called "mpg" of "No" values the same length as the number of rows in the dataframe assembleData
  2) Loop over the assembleDat object ... identifying index values of duplicate genes, updating mpg vector
  3) Merge mpg vector to assemblDat object with setting column name as "MPG"


```{r}
mpg <- rep('No', nrow(assembleDat))

for (i in unique(assembleDat[,"gene_name"])){
    if (is.na(i)){next}
    idxs <- which(assembleDat[,"gene_name"] == i)
    max_idx <- which(assembleDat$Avg[idxs] == max(assembleDat$Avg[idxs]))[1]
    
    mpg[idxs[max_idx]] <- 'Yes'
}

assembleDat$MPG <- mpg
```

### Additional boxplots after data filtering

This plot uses the annotation and expression level calculations to select only expressed protein coding genes

```{r, fig.width=10,fig.height=6}
l2cpm.select <- assembleDat %>% filter(MPG == "Yes" &
                                   gene_biotype == "protein_coding" &
                                   LowExpLowVar == "No") %>% select(ends_with(".l2cpm"))


dim(l2cpm.select)

ggplot(stack(as.data.frame(l2cpm.select)), aes(x = ind, y = values)) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ggtitle("l2cpm Expressed Protein coding boxplot") 
```

## Differential Expression Testing

### Convert columns to factors

#### This utility can set up a string for setting levels using data in the samples object

The order of levels may not be what is desired but the string makes it easy to adjust.

```{r}
# Create a vector of unique Mouse labels formatted for factor levels
Condition_levels <- samples %>%
  distinct(Condition) %>%
  arrange(Condition) %>%
  pull(Condition)

# Reformat the vector to a string that can be used to set level order
Condition_levels_string <- paste0("c(", paste(shQuote(Condition_levels, type = "cmd"), collapse = ","), ")")
cat(Condition_levels_string)
```

#### Setting factors and orders of levels

```{r}
samples$Condition <- factor(samples$Condition, levels = c("AL.WT", "AL.KO", "F.WT", "F.KO", "Rf2h.WT", "Rf2h.KO"))
samples$Condition2 <- factor(samples$Condition2, levels = c("AL.WT_stem", "AL.KO_stem", "F.WT_stem", "F.KO_stem", "Rf2h.WT_stem", "Rf2h.KO_stem"))
samples$Diet <- factor(samples$Diet, levels = c("AL","F", "Rf2h"))
samples$Odc1_status <- factor(samples$Odc1_status, levels = c("WT","KO"))
samples$Gender <- factor(samples$Gender, levels = c("F","M"))
```

### Create dds object from tximport data setting one of the conditions as reference

```{r}
dds <- DESeqDataSetFromMatrix(countData = intCt,
                              colData = samples,
                              design = ~ Condition)

dds$Condition <- relevel(dds$Condition, "AL.WT")
```

## Prepare an exploratory PCA plots

By default, plotPCA using the 500 most variable genes for PCA 1 vs PCA 2 plots
this can be adjusted with ntop argument

```{r,fig.width=7,fig.height=7}
vsd <- vst(dds, blind=TRUE)

plotPCA(vsd, intgroup="NameForProcessing", ntop=500) + geom_point(size = 4) + ggtitle("NameForProcessing")
plotPCA(vsd, intgroup="Condition", ntop=500) + geom_point(size = 4) + ggtitle("Condition")
plotPCA(vsd, intgroup="Condition2", ntop=500) + geom_point(size = 4) + ggtitle("Condition2")
plotPCA(vsd, intgroup="Replicate", ntop=500) + geom_point(size = 4) + ggtitle("Replicate")
plotPCA(vsd, intgroup="Cell", ntop=500) + geom_point(size = 4) + ggtitle("Cell")
plotPCA(vsd, intgroup="Diet", ntop=500) + geom_point(size = 4) + ggtitle("Diet")
plotPCA(vsd, intgroup="Odc1_status", ntop=500) + geom_point(size = 4) + ggtitle("Odc1_status")
plotPCA(vsd, intgroup="Gender", ntop=500) + geom_point(size = 4) + ggtitle("Gender")
plotPCA(vsd, intgroup="ReadCount", ntop=500) + geom_point(size = 4) + ggtitle("ReadCount")
plotPCA(vsd, intgroup="MapRate", ntop=500) + geom_point(size = 4) + ggtitle("Mapping Rate")


pcaData <- plotPCA(vsd, intgroup="Condition", returnData=TRUE, ntop=500)
pcaData
#write.xlsx(pcaData, colNames=TRUE, rowNames=TRUE, overwrite = TRUE, file="pca.xlsx")
```

### Prepare PCA plot figure

this code needs to be updated to automatically set x and y scales

```{r,fig.height=12,fig.width=14}
# pcaFig <- plotPCA(vsd, intgroup="Condition2", ntop=500) + 
#   geom_point(size = 4) +
#   geom_label_repel(aes(label = Condition2), vjust = -0.5, hjust = 0.5, 
#                    force = 2, min.segment.length = 0.1, max.overlaps = 100) +
#   scale_y_continuous(limits = c(-16, 16), breaks = seq(-16, 16, by = 2)) +
#   scale_x_continuous(limits = c(-22, 22), breaks = seq(-22, 22, by = 2)) +
#   ggtitle("Condition2")
# pcaFig
# 
# pdf('prog_pcaFig.pdf',width=12, height=14)
# pcaFig
# dev.off()
```

# Check counts for some genes of interest and gender genes

Adjust the list as needed
```{r, fig.width=10,fig.height=6}
genes.To.check <- c("Lgr5", "Odc1", "Ddx3y", "Xist")

# Define the desired order for Condition
condition_levels <- c("AL.WT", "AL.KO",  
                      "F.WT", "F.KO", 
                      "Rf2h.WT", "Rf2h.KO")

for (gene in genes.To.check) {
  ens.ID <- gene2typesym %>% filter(gene_name == gene)
  
  if (nrow(ens.ID) == 0) {
    warning(paste("Gene", gene, "not found in gene2typesym"))
    next
  }

  cts.ToPlot <- plotCounts(dds, gene = ens.ID$gene_id, intgroup = c("Condition", "Cell"), returnData = TRUE)
  
  # Apply the correct factor levels to Condition2
  cts.ToPlot$Condition2 <- factor(cts.ToPlot$Condition, levels = condition_levels)

  p <- ggplot(cts.ToPlot, aes(x = Condition2, y = count, color = Cell)) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.9) +
    labs(title = paste("Normalized counts for", gene), y = "Normalized count") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p)
}
```

# Differential Expression Comparisons

## There are 15 distinct comparisons:

### Reference: AL.WT

-   AL.KO_vs_AL.WT\
-   F.WT_vs_AL.WT\
-   F.KO_vs_AL.WT\
-   Rf2h.WT_vs_AL.WT\
-   Rf2h.KO_vs_AL.WT

### Reference: F.WT

-   AL.KO_vs_F.WT\
-   F.KO_vs_F.WT\
-   Rf2h.WT_vs_F.WT\
-   Rf2h.KO_vs_F.WT

### Reference: Rf2h.WT

-   AL.KO_vs_Rf2h.WT\
-   F.KO_vs_Rf2h.W\
-   Rf2h.KO_vs_Rf2h.WT

### Reference: AL.KO

-   F.KO_vs_AL.KO\
-   Rf2h.KO_vs_AL.KO

### Reference: F.KO

-   Rf2h.KO_vs_F.KO

There is a gender imbalance in many of the conditions that is apparent from PCA plots. As a result, gender will be accounted for in the differential expression testing

## Re-level to AL.WT_stem

```{r}
dds <- DESeqDataSetFromMatrix(countData = intCt,
                       colData=samples,
                       design = ~ Gender + Condition)

dds$Condition <- relevel(dds$Condition, "AL.WT")
```

### Run tests vs AL.WT

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs AL.WT_stem comparisons coef

```{r}
vs_AL.WT.comparisons <- resultsNames(dds)[c(3,4,5,6,7)]
vs_AL.WT.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_AL.WT.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to F.WT

```{r}
dds$Condition <- relevel(dds$Condition, "F.WT")
```

### Run tests vs F.WT

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs F.WT comparisons coef

### Reference: F.WT

```{r}
vs_F.WT.comparisons <- resultsNames(dds)[c(4,5,6,7)]
vs_F.WT.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_F.WT.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to Rf2h.WT_stem

```{r}
dds$Condition <- relevel(dds$Condition, "Rf2h.WT")
```

### Run tests vs Rf2h.WT

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs Rf2h.WT_stem comparisons coef

```{r}
vs_Rf2h.WT.comparisons <- resultsNames(dds)[c(5,6,7)]
vs_Rf2h.WT.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_Rf2h.WT.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to AL.KO_stem

```{r}
dds$Condition <- relevel(dds$Condition, "AL.KO")
```

### Run tests vs AL.KO

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs AL.KO comparisons coef

```{r}
vs_AL.KO.comparisons <- resultsNames(dds)[c(6,7)]
vs_AL.KO.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_AL.KO.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Re-level to F.KO

```{r}
dds$Condition <- relevel(dds$Condition, "F.KO")
```

### Run tests vs F.KO

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

### vs F.KO comparisons coef

```{r}
vs_F.KO.comparisons <- resultsNames(dds)[c(7)]
vs_F.KO.comparisons
```

### Prepare results and join to data table

```{r}
for (i in vs_F.KO.comparisons){
  res <- as.data.frame(lfcShrink(dds, coef=i, type="apeglm"))
  compar_name <- i
  cleanCompar <- str_replace(compar_name, "Condition_", "")
  stat_col <- paste('WaldStatistic', compar_name, sep = '_')
  stat <- dds@rowRanges@elementMetadata[,stat_col]
  res$stat <- stat

  res <- res %>% select(c("log2FoldChange", "pvalue", "padj","stat"))
  res <- res %>% mutate(absFC = abs(log2FoldChange))
  res <- res %>% dplyr::mutate(negLogP = -log10(padj))

  colnames(res) <- paste(cleanCompar, colnames(res), sep = '.')
  res <- tibble::rownames_to_column(res, "RowName")
  assembleDat <- merge(x=assembleDat,y=res, by.x=c("Row.names"),by.y=c("RowName"), all=TRUE)
}
```

## Write output file 

```{r}
write.xlsx(assembleDat,colNames=TRUE, quote=FALSE, rowNames=FALSE, file="250430Yil_stem_assembled.xlsx")
```

# Comparison list for checking

```{r}
comparisons.check <- as.data.frame(c(vs_AL.WT.comparisons,
                                     vs_F.WT.comparisons,
                                     vs_Rf2h.WT.comparisons,
                                     vs_AL.KO.comparisons,
                                     vs_F.KO.comparisons))

colnames(comparisons.check) <- "comparisons"

comparisons.check
```

## Upset Plots

```{r fig.width=8,fig.height=6}
# Filter the data frame based on your initial criteria
filtered_data <- assembleDat %>% filter(MPG == "Yes" & gene_biotype == "protein_coding")

# Get the column names ending with .padj and .absFC
padj_columns <- names(filtered_data)[str_detect(names(filtered_data), "\\.padj$")]
absfc_columns <- names(filtered_data)[str_detect(names(filtered_data), "\\.absFC$")]

# Extract prefixes for naming the lists
prefixes <- str_remove(padj_columns, "\\.padj$")

# Create a list of data frames, each containing genes for a specific .padj column and .absFC column
gene_lists <- map2(padj_columns, prefixes, ~ {
  list_name <- .y
  fc_column <- paste0(list_name, ".absFC")
  genes <- filtered_data %>%
    filter(.data[[.x]] < 0.05 & .data[[fc_column]] > 1) %>%
    select(gene_name) %>%
    pull(gene_name)
  setNames(list(genes), list_name)
})

# Flatten the list of lists
gene_lists <- flatten(gene_lists)

# Select a substet of the lists

gene_lists_subset <- gene_lists[c(1,7,12)]
names(gene_lists_subset)

# Convert the list of genes into a data frame suitable for UpSet plot
gene_lists_binary <- fromList(gene_lists_subset)

# Generate the UpSet plot
UpSetR::upset(gene_lists_binary, sets = names(gene_lists_subset), order.by = "freq",text.scale = 2,show.numbers = "yes")
```

## Volcano plots

### Prepare gene expression data for plotting

```{r}
vol_df <- assembleDat %>% dplyr::filter(MPG == "Yes" &
                                    gene_biotype == "protein_coding" &
                                    LowExpLowVar == "No") %>% dplyr::select(gene_name, 
                                                                            ends_with(".log2FoldChange"),
                                                                            ends_with(".padj"),
                                                                            ends_with(".absFC"))
```

## Adjust the data table for better plotting

### Replaces NA value padj columns with 1 and NA value log2FoldChange with 0 and sets y axis according to max(abs(LFC))

```{r}
vol_df.edit <- vol_df

vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(is.na(.), 1, .))
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".log2FoldChange")), ~ ifelse(is.na(.), 0, .))
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".absFC")), ~ ifelse(is.na(.), 0, .))
```

### joins list of annotated genes for coloring

gene lists of interest

```{r}
polyamine <- as.data.frame(c("Odc1","Amd1","Srm","Sms","Sat1","Smox","Paox"))
colnames(polyamine) <- "gene_name"
polyamine <- polyamine %>% dplyr::mutate(Polyamine = gene_name)

fasting <- as.data.frame(c("Acaa1a","Acaa1b","Acad8","Acadm","Acot2","Acox1","Acsl3",
                           "Angptl4","Cpt1a","Crot","Decr1","Decr2","Dgat1","Ech1","Etfb",
                           "Fabp1","Fabp2","Hoxa1","Hadhb","Hmgcl","Hmgcs2","Jag2","Pdk4","Ucp2"))
colnames(fasting) <- "gene_name"
fasting <- fasting %>% dplyr::mutate(Fasting = gene_name)

Rf.activated.Hsp <- assembleDat %>% dplyr::filter(gene_biotype == "protein_coding",
                                                  MPG == "Yes",
                                                  str_starts(gene_name, "Hsp"),
                                                  Rf2h.WT_vs_AL.WT.log2FoldChange > 0.75,
                                                  Rf2h.WT_vs_AL.WT.padj < 0.05) %>% dplyr::select(gene_name)

Rf.activated.Hsp <- Rf.activated.Hsp %>% dplyr::mutate(Rf.activated.Hsp = gene_name)

vol_df.edit <- merge(vol_df.edit,polyamine,by.x="gene_name",by.y="gene_name", all.x=TRUE, all.y = FALSE)
vol_df.edit <- merge(vol_df.edit,fasting,by.x="gene_name",by.y="gene_name", all.x=TRUE, all.y = FALSE)
vol_df.edit <- merge(vol_df.edit,Rf.activated.Hsp,by.x="gene_name",by.y="gene_name", all.x=TRUE, all.y = FALSE)
```

### applies a ceiling to -log10(adp)

This may not be necessary but can help when there are a handful of genes with large -log10(adp)

```{r}
vol_df.edit <- vol_df.edit <- vol_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(. < 1e-10, 1e-10, .))
```

### sets y axis according to max(abs(LFC))

```{r}
maxFCs <- vol_df.edit %>% summarize(across(ends_with(".log2FoldChange"), ~ max(abs(.), na.rm = TRUE)))
maxFCs
max(maxFCs)
```

### Prepare vector of plot comparisons

```{r}
comparisons_to_plot <- colnames(vol_df.edit)
comparisons_to_plot <- grep("log2FoldChange$", comparisons_to_plot, value = TRUE)
comparisons_to_plot <- sub("\\.log2FoldChange$", "", comparisons_to_plot)
comparisons_to_plot
```

### Plotting the Comparisons - top n fold changes

```{r,fig.width=10,fig.height=10}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  vol_df.edit <- vol_df.edit %>%
    mutate(
      toColor = if_else(vol_df.edit[[col_absFC]] >= 1 & vol_df.edit[[col_padj]] <= 0.05, "Gene", "Empty")
    )

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)
  
  # Create sorted_df so that labels are drawn on top
  # By pvalue
  # sorted_df <- vol_df.edit %>% arrange(vol_df.edit[[col_padj]])
  # By foldChange
  sorted_df <- vol_df.edit %>% arrange(desc(vol_df.edit[[col_absFC]]))

  # Calculate the top n column - adjust the number as desired - these will be labeled with symbol
  sorted_df <- sorted_df %>%
    mutate(top = if_else(row_number() <= 25, gene_name, ""))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for", i)
  p <- ggplot(sorted_df, aes(x = -log10(.data[[col_padj]]), y = .data[[col_log2FoldChange]])) +
    geom_point(aes(color = toColor), size = 2) +
    geom_label_repel(data = filter(sorted_df, !is.na(top)), aes(label = top), box.padding = 0.5, size = 4,max.overlaps=50) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-1, maxFC+1)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)
  print(p)
}
```


### Plotting the Comparisons - polyamine focus

```{r,fig.width=8,fig.height=8}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)

  # Separate data frames for points and labels
  points_df <- sorted_df %>% filter(is.na(Polyamine))
  labels_df <- sorted_df %>% filter(!is.na(Polyamine))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for Polyamine genes - ", i)
  p <- ggplot() +
    geom_point(data = points_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "grey70"), size = 2) +
    geom_point(data = labels_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "red"), size = 3) +
    geom_label_repel(data = labels_df, aes(x = -log10(.data[[col_padj]]),
                                           y = .data[[col_log2FoldChange]],
                                           label = Polyamine), box.padding = 0.5, size = 3, max.overlaps = 100) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-0.5, maxFC+0.5)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)

  print(p)
}
```

### Plotting the Comparisons - Fasting focus

```{r,fig.width=8,fig.height=8}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)

  # Separate data frames for points and labels
  points_df <- sorted_df %>% filter(is.na(Fasting))
  labels_df <- sorted_df %>% filter(!is.na(Fasting))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for fasting genes - ", i)
  p <- ggplot() +
    geom_point(data = points_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "grey70"), size = 2) +
    geom_point(data = labels_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "red"), size = 3) +
    geom_label_repel(data = labels_df, aes(x = -log10(.data[[col_padj]]),
                                           y = .data[[col_log2FoldChange]],
                                           label = Fasting), box.padding = 0.5, size = 3, max.overlaps = 100) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-0.5, maxFC+0.5)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)

  print(p)
}
```

### Plotting the Comparisons - Rf activated Hsp focus

```{r,,fig.width=8,fig.height=8}
for (i in comparisons_to_plot) {
  # Highlight DEGs
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  # Identify the maximum fold change for each comparison
  maxFC <- round(max(vol_df.edit[[col_absFC]]), 0)

  # Identify the maximum -log10 pvalue for each comparison
  maxP <- round(max(-log10(vol_df.edit[[col_padj]])), 0)

  # Separate data frames for points and labels
  points_df <- sorted_df %>% filter(is.na(Rf.activated.Hsp))
  labels_df <- sorted_df %>% filter(!is.na(Rf.activated.Hsp))

  # Create the volcano plot
  plot_title <- paste("Volcano Plot for Rf.activated.Hsp genes - ", i)
  p <- ggplot() +
    geom_point(data = points_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "grey70"), size = 2) +
    geom_point(data = labels_df, aes(x = -log10(.data[[col_padj]]), 
                                     y = .data[[col_log2FoldChange]], color = "red"), size = 3) +
    geom_label_repel(data = labels_df, aes(x = -log10(.data[[col_padj]]),
                                           y = .data[[col_log2FoldChange]],
                                           label = Rf.activated.Hsp), box.padding = 0.5, size = 3, max.overlaps = 100) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    geom_hline(yintercept = 0, color = "blue") +
    theme(
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(colour = "black"),
      legend.position = "none"
    ) +
    scale_y_continuous(breaks = seq(-maxFC, maxFC, 1), limits = c(-maxFC-0.5, maxFC+0.5)) +
    scale_x_continuous(limits = c(0, maxP+1)) +
    ggtitle(plot_title)

  print(p)
}
```

# Check counts for some differentially expressed genes

Adjust the list as needed

```{r}
# Ensure Condition is a factor in the DESeqDataSet
dds$Condition <- factor(dds$Condition, levels = c("AL.WT", "AL.KO", "F.WT", "F.KO", "Rf2h.WT", "Rf2h.KO"))

genes.To.check <- c("Cpt1a", "Fabp1", "Fabp2", "Decr1", "Hmgcs2", "Pdk4")

for (gene in genes.To.check) {
  ens.ID <- gene2typesym %>% filter(gene_name == gene)
  
  if (nrow(ens.ID) == 0) {
    warning(paste("Gene", gene, "not found in gene2typesym"))
    next
  }

  cts.ToPlot <- plotCounts(dds, gene = ens.ID$gene_id, intgroup = "Condition", returnData = TRUE)
  print(cts.ToPlot)
  
  cts.ToPlot$Condition <- factor(cts.ToPlot$Condition,
                                 levels = c("AL.WT", "AL.KO", "F.WT", "F.KO", "Rf2h.WT", "Rf2h.KO"))

  p <- ggplot(cts.ToPlot, aes(x = Condition, y = count)) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.9, color = "brown4") +
    labs(title = paste("Normalized counts for", gene), y = "Normalized count")
  
  print(p)
}
```

# Import Previously processed data and compare

```{r}
# tt <- read.xlsx("../tximport_mm/250430Yil_stem_assembled.xlsx")
# old.clean <- tt %>% dplyr::filter(gene_biotype == "protein_coding" & MPG == "Yes") %>% 
#   dplyr::select(Row.names,gene_name,ends_with(".log2FoldChange"))
# old.clean <- old.clean %>% rename_with(~ paste0("old_", .), .cols = -Row.names)
# 
# new.clean <- assembleDat %>% dplyr::select(Row.names,ends_with(".log2FoldChange"))
# new.clean <- new.clean %>% rename_with(~ paste0("new_", .), .cols = -Row.names)
# 
# together <- merge(old.clean,new.clean,by.x="Row.names",by.y="Row.names",all.x=TRUE,all.y=FALSE)
```


```{r,fig.width=10,fig.height=10}
# # Get all column names
# cols <- colnames(together)
# 
# # Identify old and new log2FoldChange columns
# old_cols <- grep("^old_.*\\.log2FoldChange$", cols, value = TRUE)
# new_cols <- grep("^new_.*\\.log2FoldChange$", cols, value = TRUE)
# 
# # Loop through matching pairs
# for (old_col in old_cols) {
#   suffix <- sub("^old_", "", old_col)
#   new_col <- paste0("new_", suffix)
#   
#   if (new_col %in% new_cols) {
#     p <- ggplot(together, aes_string(x = old_col, y = new_col)) +
#       geom_point(color = "steelblue", size = 2) +
#       theme_minimal() +
#       labs(
#         title = paste("Scatter Plot of", old_col, "vs", new_col),
#         x = old_col,
#         y = new_col
#       )
#     
#     print(p)
#   }
# }
```

# GSEA - Functional Class Sorting

## Preparing rank files for GSEA

Extract protein coding genes from assembled data files, adjust the data and write rnk file in gsea directory
this code chunk prepares rnk files for Linux GSEA runs

```{r}
system("mkdir -p ../gsea_v3 && chmod 775 ../gsea_v3")

rnkFile <- assembleDat %>% filter(MPG == "Yes" & gene_biotype == "protein_coding") %>% dplyr::select('gene_name',ends_with(".stat"))

names <- colnames(rnkFile)

for (i in names[-1]){
  d<-rnkFile %>%  dplyr::select('gene_name',all_of(i))
  names(d)[1] <- '#Gene'
  d<-d[complete.cases(d), ]
  write.table(d, sep='\t',file=paste0("../gsea_v3/",i,"_stem.rnk"),col.names=TRUE, quote=FALSE, row.names=FALSE)
}
```

# Single sample Gene Set Enrichment Analysis (ssGSEA)

## Extract and prepare gene expression matrix in .gct format from dds object

```{r}
# Extract the normalized count matrix from the dds
normalized_counts <- as.data.frame(counts(dds, normalized = TRUE))
# rename columns - this depends on 3pDGE vs standard
normalized_counts <- normalized_counts %>% rename_with(~ str_replace(., "\\.intCt$", ".normCt"), ends_with(".intCt"))
# if standard RNA-Seq just append .normCt
#normalized_counts <- rename_with(normalized_counts, ~paste0(.x, ".normCt"))

# Annotate counts table with gene information
normalized_counts <- merge(normalized_counts, gene2typesym, by.x=0, by.y="gene_id", all.x=TRUE, all.y=TRUE)

# Annotate counts table with MPG from assembleData
expInfo <- assembleDat %>% dplyr::select(Row.names,MPG)
normalized_counts <- merge(normalized_counts,expInfo, by.x="Row.names", by.y="Row.names", all.x=TRUE, all.y=TRUE)

# Calculate row sum
normalized_counts <- normalized_counts %>% mutate(Sum = rowSums(select(., ends_with(".normCt"))))

# Filter to protein coding genes, MPG = yes and Sum  > 0
normalized_counts <- normalized_counts %>% dplyr::filter(gene_biotype == "protein_coding" & MPG == "Yes" & Sum > 0)

# Reorder dataframe and select only columns needed for .gct file
normalized_counts <- normalized_counts %>% dplyr::select(gene_name,Row.names,ends_with(".normCt"))

# Reorder data columns according to levels
order.tmp <- samples %>% arrange(Condition)
colOrder <- paste0(order.tmp$Name, ".normCt")
normalized_counts <- normalized_counts %>% dplyr::select(gene_name,Row.names, all_of(colOrder))

# Rename ID columns
normalized_counts <- normalized_counts %>% dplyr::rename(NAME = gene_name, Description = Row.names)

# Prepare gct file
numDatCols <- normalized_counts %>% select(ends_with(".normCt")) %>% colnames() %>% length()
numDatRows <- normalized_counts %>% rownames() %>% length()

filename1 <- paste("250430Yil_stem_normalizedCounts", ".gct", sep="")
cat("#1.2", "\n", sep="\t", file=filename1)
cat(numDatRows, numDatCols, "\n", sep="\t", file=filename1, append = TRUE)

write.table(normalized_counts, file=filename1, row.names=FALSE, quote = FALSE, sep="\t", append=TRUE)
```

## run ssGSEA

```{r}
capture.output(
  ssGSEA.project.dataset(input.ds = "250430Yil_stem_normalizedCounts.gct", 
                         output.ds = "mh_stem_ssGSEA", 
                         gene.sets.dbfile.list = "/annotationFiles/v2025/mh.all.v2025.1.Mm.symbols.gmt"),
  file = "mh_ssgsea_stem_log.txt"
  )


capture.output(
  ssGSEA.project.dataset(input.ds = "250430Yil_stem_normalizedCounts.gct", 
                         output.ds = "m2cp_stem_ssGSEA", 
                         gene.sets.dbfile.list = "/annotationFiles/v2025/m2.cp.v2025.1.Mm.symbols.gmt"),
  file = "m2cp_ssgsea_stem_log.txt"
  )
```

## Import ssGSEA results, row-center the results and export to excel

```{r}
ssGSEA_file_list <- list.files(pattern = "_stem_ssGSEA.gct$", full.names = TRUE)

read_gct_file <- function(file) {
  read.table(file, header = TRUE, sep = "\t", skip = 2, stringsAsFactors = FALSE)
}

ssGSEA_gct_list <- lapply(ssGSEA_file_list, read_gct_file)
ssGSEA.results <- do.call(rbind, ssGSEA_gct_list)
ssGSEA.results <- ssGSEA.results %>% dplyr::select(-Description)

# filter out "combined up and down gene sets"

#ssGSEA.results <- ssGSEA.results %>% dplyr::filter(!grepl("combined UP & DN", Description))
ssGSEA.results <- ssGSEA.results %>% rename_with(~ str_replace(., "\\.normCt$", ".ssgsea"), ends_with(".normCt"))

ssGSEA.results <- ssGSEA.results %>% remove_rownames %>% column_to_rownames(var="Name")
ssgsea_df <- ssGSEA.results

# Extract group labels from column names - the text pattern here will vary a lot between experiments
sample_info <- tibble(
  Sample = colnames(ssgsea_df),
  Group = str_extract(colnames(ssgsea_df), "(AL|F|Rf2h)\\.(WT|KO)")
)

# Transpose the data so pathways are rows and samples are columns
ssgsea_df_t <- t(ssgsea_df)
colnames(ssgsea_df_t) <- rownames(ssGSEA.results)
ssgsea_df_t <- as.data.frame(ssgsea_df_t)


# Add group info to the transposed data
ssgsea_df_t$Group <- sample_info$Group
```

# Function to perform t-tests across groups for each pathway

```{r}
test_pathways <- function(data, group_col = "Group") {
  pathways <- colnames(data)[!colnames(data) %in% group_col]
  results <- lapply(pathways, function(pathway) {
    df <- data[, c(pathway, group_col)]
    colnames(df) <- c("Score", "Group")
    
    # Perform ANOVA across all groups
    aov_result <- aov(Score ~ Group, data = df)
    p_val <- summary(aov_result)[[1]][["Pr(>F)"]][1]
    
    # Return pathway name and p-value
    tibble(Pathway = pathway, P_value = p_val)
  })
  
  bind_rows(results) %>% arrange(P_value)
}
```

### Run the test to identify group-level pathway changes

```{r}
pathway_stats <- test_pathways(ssgsea_df_t)
# Select the top n pathways with most significant group-level differences
PathToPlot <- head(pathway_stats$Pathway, 40)
```

## Standardize the data

```{r}
ssGSEA.results <- ssGSEA.results %>% data.matrix()
rc.ssGSEA.results <- calculate_zscore(ssGSEA.results)
rc.ssGSEA.results.df <- as.data.frame(rc.ssGSEA.results)
rc.ssGSEA.results.df <- rc.ssGSEA.results.df %>% rename_with(~ str_replace(., "\\.ssgsea$", ".rc_ssgsea"), ends_with(".ssgsea"))

#write.xlsx(rc.ssGSEA.results.df,colNames=TRUE, quote=FALSE, rowNames=TRUE, file="rc.ssGSEA.results.xlsx")
```

## Heatmap of specific gene sets from ssGSEA

```{r,fig.width=12,fig.height=12}
rc.ssGSEA.results.select <- rc.ssGSEA.results.df %>%
  rownames_to_column("Pathway") %>%
  filter(Pathway %in% PathToPlot) %>%
  column_to_rownames("Pathway")

colnames(rc.ssGSEA.results.select) <- str_remove(colnames(rc.ssGSEA.results.select), "\\.rc_ssgsea$")

# Extract group labels from column names
group_labels <- str_extract(colnames(rc.ssGSEA.results.select), "(AL|F|Rf2h)\\.(WT|KO)")

# Ensure it's a proper factor
group_factor <- factor(group_labels, levels = unique(group_labels))

# Define colors for each group
group_colors <- structure(
  c("steelblue", "skyblue", "darkgreen", "lightgreen", "salmon", "purple"),
  names = levels(group_factor)
)

# Create column annotation (no need to name the factor)
col_ha <- HeatmapAnnotation(
  Group = group_factor,
  col = list(Group = group_colors),
  show_annotation_name = TRUE
)

Heatmap(
  rc.ssGSEA.results.select,
  name = "Z-score",
  top_annotation = col_ha,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 8),
  row_names_gp = gpar(fontsize = 8)
)
```

## Heatmaps

# Heatmaps

## Prepare gene expression data for plotting

```{r}
heat_df <- assembleDat %>% dplyr::filter(MPG == "Yes" &
                                    gene_biotype == "protein_coding" &
                                    LowExpLowVar == "No")

# Reorder data columns according to levels
order.tmp <- samples %>% arrange(Condition)
colOrder <- paste0(order.tmp$Name, ".l2cpm")

heat_df <- heat_df %>% dplyr::select(gene_name, all_of(colOrder),
                                     ends_with(".log2FoldChange"),
                                     ends_with(".padj"),
                                     ends_with(".absFC"))
```

## Adjust the data table for better plotting

### Replaces NA value padj columns with 1 and NA value log2FoldChange with 0 and sets y axis according to max(abs(LFC))

```{r}
heat_df.edit <- heat_df

heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".log2FoldChange")), ~ ifelse(is.na(.), 0, .))
heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".padj")), ~ ifelse(is.na(.), 1, .))
heat_df.edit <- heat_df.edit <- heat_df.edit %>% mutate_at(vars(ends_with(".absFC")), ~ ifelse(is.na(.), 0, .))
```

## Prepare vector of comparisons to plot

```{r}
comparisons_to_plot <- colnames(heat_df.edit)
comparisons_to_plot <- grep("log2FoldChange$", comparisons_to_plot, value = TRUE)
comparisons_to_plot <- sub("\\.log2FoldChange$", "", comparisons_to_plot)
comparisons_to_plot
```

This extracts l2cpm data for some of the most differential expressed
protein coding genes Adjust filters and check row counts

### Plotting the Comparisons - top n fold changes

```{r,fig.width=13,fig.height=10}
for (i in comparisons_to_plot) {
  col_log2FoldChange <- paste0(i, ".log2FoldChange")
  col_padj <- paste0(i, ".padj")
  col_absFC <- paste0(i, ".absFC")

  heat_df.toPlot <- heat_df.edit %>%
    filter(.data[[col_absFC]] >= 1 & .data[[col_padj]] <= 0.05) %>%
    slice_max(order_by = .data[[col_absFC]], n = 40) %>%
    select(gene_name, ends_with(".l2cpm"))

  # Skip if no rows to plot
  if (nrow(heat_df.toPlot) == 0) {
    message(paste("No DEGs found for", i))
    next
  }

  # Set gene_name as rownames and rename columns
  rownames(heat_df.toPlot) <- heat_df.toPlot$gene_name
  heat_df.toPlot <- heat_df.toPlot %>% select(-gene_name)
  heat_df.toPlot <- heat_df.toPlot %>% rename_with(~ str_remove(., "\\.l2cpm$"))

  # Convert to matrix
  heat_matrix <- as.matrix(heat_df.toPlot)

  # Standardize the data
  heat_matrix.zscore <- calculate_zscore(heat_matrix)
  
  # top annotation
  # Extract group labels from column names
  group_labels <- str_extract(colnames(heat_matrix.zscore), "(AL|F|Rf2h)\\.(WT|KO)")
  
  # Ensure it's a proper factor
  group_factor <- factor(group_labels, levels = unique(group_labels))
  
  # Define colors for each group
  group_colors <- structure(
    c("steelblue", "skyblue", "darkgreen", "lightgreen", "salmon", "purple"),
    names = levels(group_factor)
  )
  
  # Create column annotation (no need to name the factor)
  col_ha <- HeatmapAnnotation(
    Group = group_factor,
    col = list(Group = group_colors),
    show_annotation_name = TRUE
  )

  # Create the heatmap
  p <- Heatmap(
    heat_matrix.zscore,
    top_annotation = col_ha,
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    row_names_side = "left",
    column_names_rot = 90,
    column_names_side = "top",
    heatmap_legend_param = list(title = paste0(i, " DEG heatmap"))
  )
  print(p)
}
```

# Supplemental data and GEO files

```{r}
system("mkdir -p ../geo_submission && chmod 775 ../geo_submission")
IntCt.geo <- assembleDat %>% dplyr::select(Row.names,ends_with(".intCt"))
IntCt.geo <- IntCt.geo %>% dplyr::rename(ENSID = Row.names)
IntCt.geo <- IntCt.geo %>% dplyr::rename_all(~str_replace(., "\\.intCt", ""))
write.table(IntCt.geo,file="../geo_submission/250430Yil_stem_intCt.txt", sep = "\t", quote = FALSE, row.names = FALSE)

l2cpm.geo <- assembleDat %>% dplyr::select(Row.names,ends_with(".l2cpm"))
l2cpm.geo <- l2cpm.geo %>% dplyr::rename(ENSID = Row.names)
l2cpm.geo <- l2cpm.geo %>% dplyr::rename_all(~str_replace(., "\\.l2cpm", ""))
write.table(l2cpm.geo,file="../geo_submission/250430Yil_stem_l2cpm.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```

# write session info

Capturing information about the R session helps document your work

```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), "250430Yil_stem_sessionInfo.txt")
```